### 2020-09-16 paper/main.tex
Paper!

\documentclass[11pt,a4paper,notitlepage]{article}
\usepackage{amssymb}
\usepackage{amssymb}
\usepackage{geometry} 
\usepackage{float}
\usepackage{booktabs}
\usepackage{algorithm,algorithmic}
\usepackage{amsmath}
\usepackage{pdflscape}
\usepackage{url}
\usepackage{natbib}
\usepackage{graphicx,color}
\usepackage[utf8]{inputenc}
\usepackage{listings} % for inline code

\allowdisplaybreaks
\geometry{a4paper,left=1in,right=1in,top=1in,bottom=1in}

\newcommand{\ft}{\color{blue}}  % to show changes in revised version (F.T.)

\begin{document}

\title{Benchmark!!1}
\author{
Fabien Tricoire$^{1}$ \\[1ex]
 \small $^1$Institute for Transport and Logistics Management,
 Vienna University of Economics and Business\\
 \small Welthandelsplatz 1, 1020 Vienna, Austria\\
  \small \texttt{fabien.tricoire@wu.ac.at}\\[2ex]
}
\date{}
\maketitle

\begin{abstract}
  bla bla bla
\end{abstract}

\section{Introduction}
\label{sec:intro}
Purposes: determine which language we can tell our students to
use to implement routing optimisation routines. Also determine which
language we can use ourselves.

Philosophy: not necessarily the best
implementation in the world, but a reasonably good one, i.e. one we
can expect from an average student. Similar algorithm in various languages.

What we measure: run time of the algorithm only. Everything else
(reading data, building initial solution) is not measured. Currently
Julia also measures compilation time, needs to be fixed.

We restrict ourselves to variable-size vectors, since they are
necessary in most algorithms for vehicle routing, for instance to
insert or delete a vertex from a tour. This is the reason why we do
not provide a C implementation. It is of course possible to
implement variable-size vectors using fixed-size arrays and
bookkeeping; however this stands directly against our stated purposes
and philosophy. Moreover, implementing a bug-free, efficient version
of variable-size vectors is unlikely to result in something better
than existing libraries.

\section{Languages and implementations considered}
\subsection{C++}
\subsection{Python}
\subsection{Java}
\subsection{Julia}
\subsection{Rust}
\subsection{JavaScript}
JavaScript is not expected to perform as well as, say, C++. However
there are several reasons why it can be appealing to implement routing
algorithms in JavaScript:
\begin{itemize}
\item Any program written in JavaScript can run on virtually any
  computer with a modern web browser, including smartphones.
\item Integrating JavaScript in a web interface is especially easy.
\item JavaScript engines have received considerable attention from
  major companies and are the subject of fierce competition on
  performance. As a result they have seen vast performance
  improvements over the years and the trend is likely to continue.
\end{itemize}

We use Node.js, which allows to run JavaScript programs from the
command line. This means that there is in fact not even the need for a
web browser. Node.js uses Google Chrome v8's JavaScript engine~\cite{nodejs}.

\section{Considered benchmarks and their format}

\section{Considerations on how to implement a distance matrix}
Perhaps the most straightforward way to implement a distance matrix in
a number of languages is to use nested arrays, i.e. each element of the
main array is an array representing a row of the distance matrix. This
is typically done in C++, Java, Python (although the structure is
officially called a \emph{list} and not an array), Javascript. C and
C++ use pointers to achieve that effect. All these languages use the
same C syntax for looking up values in the distance matrix: the
distance between vertices $i$ and $j$ using distance matrix $d$ is
written \lstinline{d[i][j]}.

Another easy way to implement a distance matrix is to use what we
call from here on a \emph{flat} distance matrix representation, which
is a single-dimensional array containing all distance values. Assuming
indices start at 0, the distance from $i$ to $j$ in array $d$ can be
coded as \lstinline{d[i*n+j]}, where $n$ is the total number of
vertices. This representation guarantees that the whole distance
matrix is stored in contiguous memory. Additionally, looking up in
arrays also takes time, so one lookup is better than two. The drawback
is that we have to pay a multiplication and an addition for each
lookup, and that lookups have to be wrapped in a function call for the
sake of readability. Such function calls can usually be inlined,
i.e. the function content is substituted to the function call, so
there should be no runtime penalty from using a function call. In
general we expect a speedup from using a flat representation.

There are two cases where the above considerations on using a flat
representation do not apply: Python does not allow inlining and Julia
has native multi-dimensional arrays. This is discussed later.

\section{Language-specific notes}
\subsection{Python}
Python does not allow to inline functions. Since using a flat matrix
requires to wrap matrix value lookup in a function, using a flat distance
matrix in Python is not beneficial. It would of course be possible to
inline by hand, i.e. formulate the correct 1-dimensional index in the
matrix for every lookup, but that would be impractical and defeat the
purposes and philosophy stated in Section~\ref{sec:intro}.
Running Python code that uses a flat representation with a JIT compiler
(e.g. Pypy, Numba) introduces automatic inlining and is likely to
remedy this.
\subsection{Julia}
Julia has native multi-dimensional arrays, which may be more efficient
than a flat representation. We will determine which version to use
based on experiment. Since Julia's arrays start with index 1, the code
for flat representation lookup of the distance between vertices $i$
and $j$ in matrix $d$ is \lstinline{d[i*n-n+j]}. It is worth noting
that it involves one subtraction on top of the addition and
multiplication used for 0-indexed languages.
\subsection{JavaScript}
All benchmarks are repetitive in nature, as they successively perform similar
operations with different input data. In JavaScript, when our
implementation loads 40 instances and performs the same
benchmarking operations on all 40 them, we observe a significant performance
hit after a few instances (2-5 times slower). In order to remedy this,
we re-start the program individually for each instance. Advanced
knowledge of this specific JavaScript engine might allow to remove that
performance hit, but this is clearly outside the scope of this
study. However it is worth noting that this can be a concern in general.

\bibliographystyle{plain}
\bibliography{benchmark}

\end{document}

### 2020-09-16 python-flat-matrix/tspbenchmark.py
new implementation: python-flat-matrix

@@ -1,0 +1,1 @@
+../python/tspbenchmark.py

### 2020-09-16 python-flat-matrix/tspdata.py

@@ -1,0 +1,45 @@
+import sys
+import copy
+import random
+import math
+import functools
+
+mapsize = 100
+
+# encapsulate TSP data: instance + starting solutions to optimise for
+# benchmarking purpose
+class TSPData:
+    def __init__(self, n, d):
+        self.n = n
+        self.dist = functools.reduce(lambda x, y: x+y, d)
+
+    def d(self, i, j):
+        return self.dist[i*self.n+j]
+        
+    def matrixstring(self):
+        rows = [ ' '.join(str(x) for x in row) for row in self.d ]
+        return '\n'.join(rows)
+    
+    def genrandom(self, n):
+        x = [ random.choice(range(mapsize)) for x in range(n) ]
+        y = [ random.choice(range(mapsize)) for x in range(n) ]
+        self.n = n
+        # Euclidean distances
+        self.d = [ [ int(math.hypot(x[i]-x[j], y[i]-y[j]) * 100)
+                     for j in range(n) ] for i in range(n) ]
+        # triangle inequality correction
+        for i in range(n):
+            for j in range(n):
+                if i == j: continue
+                for k in range(n):
+                    if i == k or j == k: continue
+                    if self.d[i][j] + self.d[j][k] < self.d[i][k]:
+                        self.d[i][k] = self.d[i][j] + self.d[j][k]
+
+
+    def genrandomcycle(self):
+        # solution generation: random permutations
+        # solutions always start and end with zero
+        points = [x for x in range(1, self.n)]
+        random.shuffle(points)
+        return [0] + points + [0]

### 2020-09-16 python-flat-matrix/tspsolution.py

@@ -1,0 +1,65 @@
+import tspdata
+import copy
+from itertools import chain
+
+class TSPSolution:
+    def __init__(self, data, permutation):
+        self.data = data
+        self.nodes = copy.copy(permutation)
+
+    def two_opt(self):
+        t = 0
+        while self.first2eimprovement():
+            t += 1
+        return t
+
+    def first2eimprovement(self):
+        # actually speeds things up!
+        tour, d = self.nodes, self.data.d
+        for p1 in range(len(tour) - 3):
+            for p2 in range(p1+2, len(tour) - 1):
+                if d(tour[p1], tour[p1+1]) + d(tour[p2], tour[p2+1]) > \
+                   d(tour[p1], tour[p2]) + d(tour[p1+1], tour[p2+1]):
+                    # improving 2-exchange found
+                    for i in range((p2-p1+1) // 2):
+                        tour[p1+1+i], tour[p2-i] = tour[p2-i], tour[p1+1+i]
+                    return True
+        return False
+
+    def or_opt(self):
+        t = 0
+        while self.firstorimprovement():
+            t += 1
+        return t
+
+    def firstorimprovement(self):
+        # actually speeds things up!
+        tour, d = self.nodes, self.data.d
+        for i in range(1, len(tour) - 1):
+            for l in range(1, 1 + min(3, len(tour)-1-i)):
+                for p in chain(range(i-1), range(i+l, len(tour)-1)):
+                    delta = d(tour[i-1], tour[i+l]) + d(tour[p], tour[i])\
+                        + d(tour[i+l-1], tour[p+1]) - d(tour[p], tour[p+1])\
+                        - d(tour[i-1], tour[i]) - d(tour[i+l-1], tour[i+l])
+                    # perform improving move
+                    if delta < 0:
+                        # store sequence to move
+                        t = [ x for x in tour[i:i+l] ]
+                        if i < p:
+                            # shift stuff left
+                            tour[i:p-l+1] = tour[i+l:p+1]
+                            # copy sequence right of the stuff
+                            tour[p+1-l:p+1] = t[:]
+                        else:
+                            # shift stuff right
+                            tour[i-1+l:p+l:-1] = tour[i-1:p:-1]
+                            # copy sequence right of the stuff
+                            tour[p+1:p+1+l] = t[:]
+                        return True
+        return False
+    
+    def cost(self):
+        total = 0
+        for i, j in zip(self.nodes[:-1], self.nodes[1:]):
+            total += self.data.d(i, j)
+        return total

### 2020-09-30 implementations/c++-hybrid-matrix/compile.sh
lns benchmark implementations: C++ (all except static), Java (all except static), Javascript

@@ -1,9 +1,1 @@
-#!/bin/bash
-
-# call this script with the COMPILER env. variable set to the wanted compiler
-# if none is specified, g++ is used
-# example: COMPILER=clang++ ./compile.sh
-
-COMPILER=${COMPILER:-g++}
-
-$COMPILER -DCOMPILER=\"$COMPILER\" -Wall -ansi -pedantic -O3 -std=c++14 main.cpp -o tspbenchmark
+../c++/compile.sh

### 2020-09-30 implementations/c++-hybrid-matrix/main.cpp

@@ -1,103 +1,1 @@
-#include "tspdata.h"
-#include "tspsolution.h"
-
-// read input file and return corresponding data and solutions
-template<class T> pair<shared_ptr<TSPData<T> >, vector<TSPSolution<T> > >
-read_data(string fname) {
-    unsigned int n = 0;
-    T *d=NULL;
-    unsigned int ditems = 0;
-    vector<vector<int> > tours;
-    vector<TSPSolution<T> > solutions;
-    unsigned int nsols = 0;
-    ifstream infile(fname.c_str());
-    if (! infile.is_open()) {
-	cerr << "Can't open file: " << fname << endl;
-	exit(2);
-    }
-    string line;
-    while (! infile.eof()) {
-	getline(infile, line);
-	// remove comments
-	line = line.substr(0, line.find("#"));
-	auto tokens = tokenize(line);
-	if (tokens.size() == 0) {
-	    continue;
-	} else if (n == 0 and tokens.size() == 2) {
-	    n = stoi(tokens[0]);
-	    nsols = stoi(tokens[1]);
-	    d = new int[n*n];
-	} else if (ditems < n*n) {
-	    stringstream gg(line);
-	    for (unsigned int i=0; i < n; i++) {
-		gg >> d[ditems++];
-	    }
-	} else if (tours.size() < nsols) {
-	    vector<int> tour;
-	    int tmp;
-	    stringstream gg(line);
-	    for (unsigned int i=0; i < n + 1; i++) {
-		gg >> tmp;
-		tour.push_back(tmp);
-	    }
-	    tours.push_back(tour);
-	}
-    }
-    auto dptr = make_shared<TSPData<T> >(TSPData<T>(n, d));
-    for (unsigned int i=0; i < tours.size(); i++) {
-    	solutions.push_back(TSPSolution<T>(dptr, tours[i]));
-    }
-    return make_pair(dptr, solutions);
-}
-
-// benchmark all solutions in the given vector
-template<class T> pair<int, double>
-benchmark_one(vector<TSPSolution<T> > solutions, string benchmarkname) {
-    int nimpr = 0;
-    double total_ls = 0;
-    int n;
-    for (unsigned int i=0; i < solutions.size(); i++) {
-	clock_t t2 = clock();
-	if (benchmarkname == "2-opt") {
-	    n = solutions[i].two_opt();
-	} else if (benchmarkname == "Or-opt") {
[…]

### 2020-09-30 implementations/c++-hybrid-matrix/run_benchmark.sh

@@ -1,9 +1,1 @@
-#!/bin/bash
-
-dirname=`basename \`pwd\``
-
-#echo "#language,instance,n,nsolutions,n_improvements,CPU"
-for i in $1/*
-do
-    ./tspbenchmark $i $2 | sed -e "s/c++/$dirname/g"
-done
+../c++/run_benchmark.sh

### 2020-09-30 implementations/c++-hybrid-matrix/tspsolution.h

@@ -1,106 +1,1 @@
-#ifndef TSPSOLUTION_H
-#define TSPSOLUTION_H
-
-#include <vector>
-#include <memory>
-
-#include "tspdata.h"
-
-template <class T> class TSPSolution {
-protected:
-    shared_ptr<TSPData<T> > data_;
-    vector<int> nodes_;
-    
-    bool first2eimprovement() {
-	for (unsigned int p1=0; p1 < nodes_.size() - 3; p1++) {
-	    for (unsigned int p2=p1+2; p2 < nodes_.size() - 1; p2++) {
-		if (data_->d(nodes_[p1], nodes_[p1+1]) +
-		    data_->d(nodes_[p2], nodes_[p2+1]) > 
-		    data_->d(nodes_[p1], nodes_[p2]) +
-		    data_->d(nodes_[p1+1], nodes_[p2+1])) {
-		    int t;
-		    for (unsigned int i=0; i < ((p2-p1+1) / 2); i++) {
-			t = nodes_[p1+1+i];
-			nodes_[p1+1+i] = nodes_[p2-i];
-			nodes_[p2-i] = t;
-		    }
-		    return true;
-		}
-	    }
-	}
-	return false;
-    }
-
-    // used in Or-opt
-    inline int or_delta(unsigned int i, unsigned int l, unsigned int p) {
-	return data_->d(nodes_[i-1], nodes_[i+l])
-	    + data_->d(nodes_[p], nodes_[i])
-	    + data_->d(nodes_[i+l-1], nodes_[p+1])
-	    - data_->d(nodes_[p], nodes_[p+1])
-	    - data_->d(nodes_[i-1], nodes_[i])
-	    - data_->d(nodes_[i+l-1], nodes_[i+l]);
-    }
-    
-    bool firstorimprovement() {
-	for (unsigned int i=1; i < nodes_.size() - 1; i++) {
-	    for (unsigned int l=1;
-		 l < 1 + min((long unsigned)3, nodes_.size() - 1 - i); l++) {
-		for (unsigned int p=0; p < i-1; p++) {
-		    int delta = or_delta(i, l, p);
-		    if (delta < 0) {
-			// first copy sequence we want to move
-			int *t = new int[l];
-			for(unsigned int j=0; j < l; j++) {
-			    t[j] = nodes_[i+j]; }
-			// next move around what will end up being right of it
-			for(unsigned int j=i-1; j > p; j--) {
-			    nodes_[j+l] = nodes_[j]; }
-			// finally write in the sequence being moved
[…]

### 2020-09-30 implementations/c++-nested-matrix/main.cpp

@@ -68,6 +68,8 @@ benchmark_one(vector<TSPSolution<T> > solutions, string benchmarkname) {
 	    n = solutions[i].two_opt();
 	} else if (benchmarkname == "Or-opt") {
 	    n = solutions[i].or_opt();
+	} else if (benchmarkname == "lns") {
+	    n = solutions[i].lns();
 	} else {
 	    cerr << "Unknown benchmark: " << benchmarkname << endl;
 	    exit(2);

### 2020-09-30 implementations/c++/main.cpp

@@ -62,6 +62,8 @@ benchmark_one(vector<TSPSolution<T> > solutions, string benchmarkname) {
 	    n = solutions[i].two_opt();
 	} else if (benchmarkname == "Or-opt") {
 	    n = solutions[i].or_opt();
+	} else if (benchmarkname == "lns") {
+	    n = solutions[i].lns();
 	} else {
 	    cerr << "Unknown benchmark: " << benchmarkname << endl;
 	    exit(2);

### 2020-09-30 implementations/c++/tspsolution.h

@@ -3,6 +3,7 @@
 
 #include <vector>
 #include <memory>
+#include <climits>
 
 #include "tspdata.h"
 
@@ -81,7 +82,7 @@ protected:
 	}
 	return false;
     }
-    
+
 public:
     TSPSolution(shared_ptr<TSPData<T> > data, const vector<int> &permutation) {
 	data_ = data;
@@ -100,6 +101,48 @@ public:
 	return t;
     }
 
+    int lns(unsigned int niter=10) {
+	T checksum = 0;
+	for (unsigned int iter=0; iter < niter; iter++) {
+	    // step 0: copy solution
+	    vector<int> tmp = nodes_;
+	    vector<int> unplanned;
+	    // step 1: destroy
+	    unsigned int where = 1;
+	    while (where < tmp.size() - 1) {
+		unplanned.push_back(tmp[where]);
+		tmp.erase(tmp.begin() + where);
+		where += 1;
+	    }
+	    // step 2: repair
+	    while (unplanned.size() > 0) {
+		unsigned int bestfrom=0, bestto=0;
+		// should be big enough and castable for any relevant type
+		T bestcost = INT_MAX;
+		for (unsigned int k=0; k < unplanned.size(); k++) {
+		    for (unsigned to=0; to < tmp.size() -1; to++) {
+			if (data_->d(tmp[to], unplanned[k]) +
+			    data_->d(unplanned[k], tmp[to+1]) -
+			    data_->d(tmp[to], tmp[to+1]) < bestcost) {
+			    bestcost = data_->d(tmp[to], unplanned[k]) +
+				data_->d(unplanned[k], tmp[to+1]) -
+				data_->d(tmp[to], tmp[to+1]);
+			    bestfrom = k;
+			    bestto = to;
+			}
+		    }
+		}
+		// perform best found insertion
+		tmp.insert(tmp.begin() + bestto + 1, unplanned[bestfrom]);
+		unplanned.erase(unplanned.begin() + bestfrom);
+		checksum += bestcost;
+	    }
+	    // step 3: move or not
+	    nodes_ = tmp;
+	}
+	return (int) checksum;
+    }
+    
     const shared_ptr<TSPData<T> > data(){ return data_; }
 };
 

### 2020-09-30 implementations/c++98/main.cpp

@@ -68,6 +68,8 @@ benchmark_one(const TSPData<T> &data, vector<TSPSolution<T> > solutions,
 	    n = solutions[i].two_opt(data);
 	} else if (benchmarkname == "Or-opt") {
 	    n = solutions[i].or_opt(data);
+	} else if (benchmarkname == "lns") {
+	    n = solutions[i].lns(data);
 	} else {
 	    cerr << "Unknown benchmark: " << benchmarkname << endl;
 	    exit(2);

### 2020-09-30 implementations/c++98/tspsolution.h

@@ -3,6 +3,7 @@
 
 #include <vector>
 #include <memory>
+#include <climits>
 
 #include "tspdata.h"
 
@@ -88,6 +89,49 @@ public:
 	while (firstorimprovement(data)) t++;
 	return t;
     }
+
+    int lns(const TSPData<T> &data, unsigned int niter=10) {
+	T checksum = 0;
+	for (unsigned int iter=0; iter < niter; iter++) {
+	    // step 0: copy solution
+	    vector<int> tmp = nodes_;
+	    vector<int> unplanned;
+	    // step 1: destroy
+	    unsigned int where = 1;
+	    while (where < tmp.size() - 1) {
+		unplanned.push_back(tmp[where]);
+		tmp.erase(tmp.begin() + where);
+		where += 1;
+	    }
+	    // step 2: repair
+	    while (unplanned.size() > 0) {
+		unsigned int bestfrom=0, bestto=0;
+		// should be big enough and castable for any relevant type
+		T bestcost = INT_MAX;
+		for (unsigned int k=0; k < unplanned.size(); k++) {
+		    for (unsigned to=0; to < tmp.size() -1; to++) {
+			if (data.d(tmp[to], unplanned[k]) +
+			    data.d(unplanned[k], tmp[to+1]) -
+			    data.d(tmp[to], tmp[to+1]) < bestcost) {
+			    bestcost = data.d(tmp[to], unplanned[k]) +
+				data.d(unplanned[k], tmp[to+1]) -
+				data.d(tmp[to], tmp[to+1]);
+			    bestfrom = k;
+			    bestto = to;
+			}
+		    }
+		}
+		// perform best found insertion
+		tmp.insert(tmp.begin() + bestto + 1, unplanned[bestfrom]);
+		unplanned.erase(unplanned.begin() + bestfrom);
+		checksum += bestcost;
+	    }
+	    // step 3: move or not
+	    nodes_ = tmp;
+	}
+	return (int) checksum;
+    }
+    
 };
 
 #endif

### 2020-09-30 implementations/java-nested-matrix/TSPBenchmark.java

@@ -1,1 +1,125 @@
-../java/TSPBenchmark.java
+import java.io.FileReader;
+import java.io.BufferedReader;
+import java.util.StringTokenizer;
+import java.util.ArrayList;
+import java.io.IOException;
+import java.io.File;
+
+class Pair<T1, T2>{
+    protected T1 key;
+    protected T2 value;
+    public Pair(T1 key, T2 value) {
+	this.key = key;
+	this.value = value;
+    }
+    public T1 getKey() { return key; }
+    public T2 getValue() { return value; }
+}
+
+public class TSPBenchmark{
+    
+    // load data from file
+    private static Pair<TSPData, TSPSolution[]> loadFromFile(String fname)
+	throws IOException {
+	FileReader fr = new FileReader(fname);
+	BufferedReader br = new BufferedReader(fr);
+	String s;
+	StringTokenizer st;
+	int nsols = 0;
+	int drow=0, srow=0;
+	int n=0;
+	int[] d = new int[0];;
+	ArrayList<ArrayList<Integer> > tours =
+	    new ArrayList<ArrayList<Integer> >();
+	while((s = br.readLine()) != null) {
+	    int t = s.indexOf('#');
+	    if (t != -1) s = s.substring(0, t);
+	    st = new StringTokenizer(s);
+	    if (st.countTokens() == 0) continue;
+	    else if (n == 0 && st.countTokens() == 2) {
+		n = Integer.parseInt(st.nextToken());
+		d = new int[n*n];
+		nsols = Integer.parseInt(st.nextToken());
+	    } else if (drow < n) {
+		int col = 0;
+		while (st.hasMoreTokens()) {
+		    d[n*drow+col++] = Integer.parseInt(st.nextToken());
+		}
+		drow += 1;
+	    } else if (srow < nsols) {
+		ArrayList<Integer> tour = new ArrayList<Integer>();
+		while (st.hasMoreTokens()) {
+		    tour.add(Integer.parseInt(st.nextToken()));
+		}
+		tours.add(tour);
+	    }
+	}
+	TSPData data = new TSPData(n, d);
+	TSPSolution[] solutions = new TSPSolution[nsols];
+	for (int i=0; i < nsols; i++) {
+	    solutions[i] = new TSPSolution(data, tours.get(i));
+	}
+	return new Pair<TSPData, TSPSolution[]>(data, solutions);
+    }
+    
+    private static Pair<Integer, Double> benchmark_one(TSPSolution[] solutions,
[…]

### 2020-09-30 implementations/java-nested-matrix/TSPData.java

@@ -12,4 +12,14 @@ public class TSPData {
 	n_ = n;
 	d_ = d;
     }
+    
+    public TSPData(int n, int[]d) {
+	n_ = n;
+	d_ = new int[n][n];
+	for (int i=0; i<n; i++) {
+	    for (int j=0; j<n; j++) {
+		d_[i][j] = d[i*n+j];
+	    }
+	}
+    }
 }

### 2020-09-30 implementations/java-nested-matrix/TSPSolution.java

@@ -1,90 +1,1 @@
-import java.util.ArrayList;
-import java.util.Arrays;
-
-public class TSPSolution {
-    private TSPData data_;
-    private ArrayList<Integer> nodes_;
-
-    public TSPSolution(TSPData data, ArrayList<Integer> nodes) {
-	data_ = data;
-	nodes_ = nodes;
-    }
-    
-    public int two_opt() {
-	int t = 0;
-	while (first2eimprovement()) t++;
-	return t;
-    }
-
-    private boolean first2eimprovement() {
-	for (int p1=0; p1 < nodes_.size() - 3; p1++) {
-	    for (int p2=p1+2; p2 < nodes_.size() -1; p2++) {
-		if (data_.d(nodes_.get(p1), nodes_.get(p1+1)) +
-		    data_.d(nodes_.get(p2), nodes_.get(p2+1)) > 
-		    data_.d(nodes_.get(p1), nodes_.get(p2)) +
-		    data_.d(nodes_.get(p1+1), nodes_.get(p2+1))) {
-		    // improving two-exchange found!
-		    int t;
-		    for (int i=0; i < (p2-p1+1) / 2; i++) {
-			t = nodes_.get(p1+1+i);
-			nodes_.set(p1+1+i, nodes_.get(p2-i));
-			nodes_.set(p2-i, t);
-		    }
-		    return true;
-		}
-	    }
-	}
-	return false;
-    }
-
-    public int or_opt() {
-	int t = 0;
-	while (firstorimprovement()) t++;
-	return t;
-    }
-
-    private boolean firstorimprovement() {
-	for (int i=1; i < nodes_.size() - 1; i++) {
-	    for (int l = 1; l < 1 + Math.min(3, nodes_.size() - 1 - i); l++) {
-		for (int p=0; p < i-1; p++) {
-		    if (or_delta(i, l, p) < 0) {
-			// first copy sequence we want to move
-			int [] t = new int[l];
-			for(int j=0; j < l; j++) { t[j] = nodes_.get(i+j); }
-			// next move around what will end up being right of it
-			for(int j=i-1; j > p; j--) {
-			    nodes_.set(j+l, nodes_.get(j));
-			}
-			// finally write in the sequence being moved
-			for (int j=0; j < l; j++) { nodes_.set(p+1+j, t[j]); }
-			return true;
-		    }
-		}
-		for (int p=i+l; p < nodes_.size() - 1; p++) {
[…]

### 2020-09-30 implementations/java/TSPBenchmark.java

@@ -74,6 +74,8 @@ public class TSPBenchmark{
 		n = solutions[i].two_opt();
 	    } else if (benchmarkname.equals("Or-opt")) {
 		n = solutions[i].or_opt();
+	    } else if (benchmarkname.equals("lns")) {
+		n = solutions[i].lns(10);
 	    } else {
 		System.err.println("Unknown benchmark: <" + benchmarkname +">");
 		System.exit(2);

### 2020-09-30 implementations/java/TSPSolution.java

@@ -87,4 +87,45 @@ public class TSPSolution {
 	    - data_.d(nodes_.get(i-1), nodes_.get(i))
 	    - data_.d(nodes_.get(i+l-1), nodes_.get(i+l));
     }
+
+    public int lns(int niter) {
+	int checksum = 0;
+	for (int iter=0; iter < niter; iter++) {
+	    // step 0: copy solution
+	    ArrayList<Integer> tmp = new ArrayList<Integer>(nodes_);
+	    ArrayList<Integer> unplanned = new ArrayList<Integer>();
+	    // step 1: destroy
+	    int where = 1;
+	    while (where < tmp.size() - 1) {
+		unplanned.add(tmp.get(where));
+		tmp.remove(where);
+		where += 1;
+	    }
+	    // step 2: repair
+	    while (unplanned.size() > 0) {
+		int bestfrom=0, bestto=0;
+		int bestcost = Integer.MAX_VALUE;
+		for (int k=0; k < unplanned.size(); k++) {
+		    for (int to=0; to < tmp.size() - 1; to++) {
+			int delta = data_.d(tmp.get(to), unplanned.get(k)) +
+			    data_.d(unplanned.get(k), tmp.get(to+1)) -
+			    data_.d(tmp.get(to), tmp.get(to+1));
+			if (delta < bestcost) {
+			    bestcost = delta;
+			    bestfrom = k;
+			    bestto = to;
+			}
+			
+		    }
+		}
+		// perform best found insertion
+		tmp.add(bestto + 1, unplanned.get(bestfrom));
+		unplanned.remove(bestfrom);
+		checksum += bestcost;
+	    }
+	    // step 3: move or not
+	    nodes_ = tmp;
+	}
+	return checksum;
+    }
 }

### 2020-09-30 implementations/javascript-nested-matrix/tspbenchmark.js

@@ -77,6 +77,45 @@ function TSPSolution(data, sequence) {
 	return false;
     };
 
+    // lns!
+    this.lns = function(niter=10) {
+	var checksum = 0;
+	for (var iter=0; iter < niter; iter++) {
+	    // step 0: copy solution
+	    var tmp = [...tour];
+	    var unplanned = [];
+	    // step 1: destroy
+	    var where = 1;
+	    while (where < tmp.length - 1) {
+		unplanned.push(tmp[where]);
+		tmp.splice(where, 1);
+		where += 1;
+	    }
+	    // step 2: repair
+	    while (unplanned.length > 0) {
+		var bestfrom=0, bestto=0, bestcost = Number.MAX_SAFE_INTEGER;
+		for (var k=0; k < unplanned.length; k++) {
+		    for (var to=0; to < tmp.length - 1; to++) {
+			var delta = d[tmp[to]][unplanned[k]] +
+			    d[unplanned[k]][tmp[to+1]] -
+			    d[tmp[to]][tmp[to+1]];
+			if (delta < bestcost) {
+			    bestcost = delta;
+			    bestfrom = k;
+			    bestto = to;
+			}
+		    }
+		}
+		// perform best found insertion
+		tmp.splice(bestto + 1, 0, unplanned[bestfrom]);
+		unplanned.splice(bestfrom, 1);
+		checksum += bestcost;
+	    }
+	    // step 3: move or not
+	    tour = tmp;
+	}
+	return checksum;
+    }
 }
 
 // this version requires node
@@ -122,6 +161,8 @@ function benchmarkOne(path, benchmarkname) {
 	    n = sol.twoOpt();
 	} else if (benchmarkname == 'Or-opt') {
 	    n = sol.orOpt();
+	} else if (benchmarkname == 'lns') {
+	    n = sol.lns();
 	} else {
 	    console.log('Unknown benchmark: ', benchmarkname);
 	}

### 2020-09-30 implementations/javascript/tspbenchmark.js

@@ -81,6 +81,46 @@ function TSPSolution(data, sequence) {
 	return false;
     };
 
+    // lns!
+    this.lns = function(niter=10) {
+	var checksum = 0;
+	for (var iter=0; iter < niter; iter++) {
+	    // step 0: copy solution
+	    var tmp = [...tour];
+	    var unplanned = [];
+	    // step 1: destroy
+	    var where = 1;
+	    while (where < tmp.length - 1) {
+		unplanned.push(tmp[where]);
+		tmp.splice(where, 1);
+		where += 1;
+	    }
+	    // step 2: repair
+	    while (unplanned.length > 0) {
+		var bestfrom=0, bestto=0, bestcost = Number.MAX_SAFE_INTEGER;
+		for (var k=0; k < unplanned.length; k++) {
+		    for (var to=0; to < tmp.length - 1; to++) {
+			var delta = d.d(tmp[to], unplanned[k]) +
+			    d.d(unplanned[k], tmp[to+1]) -
+			    d.d(tmp[to], tmp[to+1]);
+			if (delta < bestcost) {
+			    bestcost = delta;
+			    bestfrom = k;
+			    bestto = to;
+			}
+		    }
+		}
+		// perform best found insertion
+		tmp.splice(bestto + 1, 0, unplanned[bestfrom]);
+		unplanned.splice(bestfrom, 1);
+		checksum += bestcost;
+	    }
+	    // step 3: move or not
+	    tour = tmp;
+	}
+	return checksum;
+    }
+    
 }
 
 // this version requires node
@@ -126,6 +166,8 @@ function benchmarkOne(path, benchmarkname) {
 	    n = sol.twoOpt();
 	} else if (benchmarkname == 'Or-opt') {
 	    n = sol.orOpt();
+	} else if (benchmarkname == 'lns') {
+	    n = sol.lns();
 	} else {
 	    console.log('Unknown benchmark: ', benchmarkname);
 	}

### 2020-09-30 implementations/python/tspsolution.py

@@ -82,7 +82,7 @@ class TSPSolution:
                             bestcost = d[i][k] + d[k][j] - d[i][j]
                             bestnode, bestfro, bestto = k, fro, pos
                 # perform best found insertion
-                tmp.insert(pos+1, bestnode)
+                tmp.insert(bestto+1, bestnode)
                 del unplanned[bestfro]
                 checksum += bestcost
             # step 3: move or not (in our case always move)

### 2020-09-30 implementations/c++/tspsolution.h
cleaner c++ implementations of lns

@@ -121,12 +121,12 @@ public:
 		T bestcost = INT_MAX;
 		for (unsigned int k=0; k < unplanned.size(); k++) {
 		    for (unsigned to=0; to < tmp.size() -1; to++) {
-			if (data_->d(tmp[to], unplanned[k]) +
+			
+			T delta = data_->d(tmp[to], unplanned[k]) +
 			    data_->d(unplanned[k], tmp[to+1]) -
-			    data_->d(tmp[to], tmp[to+1]) < bestcost) {
-			    bestcost = data_->d(tmp[to], unplanned[k]) +
-				data_->d(unplanned[k], tmp[to+1]) -
-				data_->d(tmp[to], tmp[to+1]);
+			    data_->d(tmp[to], tmp[to+1]);
+			if (delta < bestcost) {
+			    bestcost = delta;
 			    bestfrom = k;
 			    bestto = to;
 			}

### 2020-09-30 implementations/c++98/tspsolution.h

@@ -110,12 +110,11 @@ public:
 		T bestcost = INT_MAX;
 		for (unsigned int k=0; k < unplanned.size(); k++) {
 		    for (unsigned to=0; to < tmp.size() -1; to++) {
-			if (data.d(tmp[to], unplanned[k]) +
+			T delta = data.d(tmp[to], unplanned[k]) +
 			    data.d(unplanned[k], tmp[to+1]) -
-			    data.d(tmp[to], tmp[to+1]) < bestcost) {
-			    bestcost = data.d(tmp[to], unplanned[k]) +
-				data.d(unplanned[k], tmp[to+1]) -
-				data.d(tmp[to], tmp[to+1]);
+			    data.d(tmp[to], tmp[to+1]);
+			if (delta < bestcost) {
+			    bestcost = delta;
 			    bestfrom = k;
 			    bestto = to;
 			}

### 2020-09-30 implementations/julia-array/TSPBenchmark/src/TSPBenchmark.jl
all julia implementations of LNS benchmark

@@ -108,30 +108,40 @@ function firstorimprovement(d::TSPData{T}, s::TSPSolution) where T
     return false
 end
 
-# old less efficient version
-
-# function firstorimprovement(d::TSPData{T}, s::TSPSolution) where T
-#     for i in 2:length(s.tour)-1, l in 1:(min(3, length(s.tour)-i)),
-#         p in [ 1:i-2 ; i+l:length(s.tour)-1 ]
-#         #
-#         δ = d.d[s.tour[i-1], s.tour[i+l]] + d.d[s.tour[p], s.tour[i]] +
-#             d.d[s.tour[i+l-1], s.tour[p+1]] -
-#             d.d[s.tour[p], s.tour[p+1]] - d.d[s.tour[i-1], s.tour[i]] -
-#             d.d[s.tour[i+l-1], s.tour[i+l]]
-#         # improvement found
-#         if δ < 0
-#             if i < p
-#                 s.tour = [ s.tour[1:i-1] ; s.tour[i+l:p] ; s.tour[i:i+l-1] ;
-#                            s.tour[p+1:end] ]
-#             else
-#                 s.tour = [ s.tour[1:p] ; s.tour[i:i+l-1] ; s.tour[p+1:i-1] ;
-#                            s.tour[i+l:end] ]
-#             end
-#             return true
-#         end
-#     end
-#     return false
-# end
+function lns(d::TSPData{T}, sol::TSPSolution, niter::Int=10) where T
+    checksum = zero(T)
+    for iter in 1:niter
+        # step 0: copy incumbent
+        tmp = copy(sol.tour)
+        unplanned = Int[]
+        # step 1: destroy
+        t = 2
+        while t < length(tmp)
+            push!(unplanned, tmp[t])
+            deleteat!(tmp, t)
+            t += 1
+        end
+        # step 2: repair
+        while length(unplanned) > 0
+            bestcost, bestfrom, bestto = typemax(T), 0, 0
+            for from in 1:length(unplanned), to in 1:length(tmp)-1
+                δ = d.d[tmp[to], unplanned[from]] +
+                    d.d[unplanned[from], tmp[to+1]] -
+                    d.d[tmp[to], tmp[to+1]]
+                if δ < bestcost
+                    bestcost, bestfrom, bestto = δ, from, to
+                end
+            end
+            # perform best found insertion
+            insert!(tmp, bestto + 1, unplanned[bestfrom])
+            deleteat!(unplanned, bestfrom)
+            checksum += bestcost
[…]

### 2020-09-30 implementations/julia-square-matrix/TSPBenchmark/src/TSPBenchmark.jl

@@ -106,6 +106,44 @@ function firstorimprovement(d::T, s::TSPSolution) where T
     return false
 end
 
+function lns(d::T, sol::TSPSolution, niter::Int=10) where T
+    # ugly! is there an elegant Julia solution without writing more functions?
+    ntype = typeof(d.d[1,1])
+    #
+    checksum = zero(ntype)
+    for iter in 1:niter
+        # step 0: copy incumbent
+        tmp = copy(sol.tour)
+        unplanned = Int[]
+        # step 1: destroy
+        t = 2
+        while t < length(tmp)
+            push!(unplanned, tmp[t])
+            deleteat!(tmp, t)
+            t += 1
+        end
+        # step 2: repair
+        while length(unplanned) > 0
+            bestcost, bestfrom, bestto = typemax(ntype), 0, 0
+            for from in 1:length(unplanned), to in 1:length(tmp)-1
+                δ = dist(d, tmp[to], unplanned[from]) +
+                    dist(d, unplanned[from], tmp[to+1]) -
+                    dist(d, tmp[to], tmp[to+1])
+                if δ < bestcost
+                    bestcost, bestfrom, bestto = δ, from, to
+                end
+            end
+            # perform best found insertion
+            insert!(tmp, bestto + 1, unplanned[bestfrom])
+            deleteat!(unplanned, bestfrom)
+            checksum += bestcost
+        end
+        # step 3: move to new incumbent
+        sol.tour = tmp
+    end
+    checksum
+end
+
 function benchmark_one(data::T, solutions::Vector{TSPSolution},
                        benchmarkname::String) where T
     totaltime = 0.0
@@ -115,6 +153,8 @@ function benchmark_one(data::T, solutions::Vector{TSPSolution},
             t = @CPUelapsed n = two_opt(data, s)
         elseif lowercase(benchmarkname) == "or-opt"
             t = @CPUelapsed n = or_opt(data, s)
+        elseif lowercase(benchmarkname) == "lns"
+            t = @CPUelapsed n = lns(data, s)
         else
             throw(ArgumentError("Unknown benchmark: $benchmarkname"))
         end

### 2020-09-30 implementations/python/tspsolution.py

@@ -63,6 +63,9 @@ class TSPSolution:
     def lns(self, niter=10):
         d = self.data.d
         checksum = 0
+
+        # print('Tour before:', self.nodes)
+        
         for iter in range(niter):
             # step 0: copy solution
             tmp = copy.copy(self.nodes)
@@ -73,19 +76,42 @@ class TSPSolution:
                 unplanned.append(tmp[where])
                 del tmp[where]
                 where += 1
+
+            # print('Tour after destruction:', tmp)
+        
             # step 2: repair
             while len(unplanned) > 0:
                 bestcost, bestnode, bestfro, bestto = math.inf, -1, -1, -1
                 for (fro, k) in enumerate(unplanned):
                     for ((pos, i), j) in zip(enumerate(tmp[:-1]), tmp[1:]):
+
+                        # print("Insertion of", k, "between", i, "and", j,
+                        #       "\tδ =", d[i][k] + d[k][j] - d[i][j])
+
+                        # print("\td[" + str(i) + "][" + str(k) + "] = "
+                        #         + str(d[i][k]))
+                        # print("\td[" + str(k) + "][" + str(j) + "] = "
+                        #         + str(d[k][j]))
+                        # print("\td[" + str(i) + "][" + str(j) + "] = "
+                        #         + str(d[i][j]))
+        
                         if d[i][k] + d[k][j] - d[i][j] < bestcost:
+
+                            # print('\tBEST')
+                            
                             bestcost = d[i][k] + d[k][j] - d[i][j]
                             bestnode, bestfro, bestto = k, fro, pos
                 # perform best found insertion
+
+                # print('Tour before insertion:', tmp)
+        
                 tmp.insert(bestto+1, bestnode)
                 del unplanned[bestfro]
                 checksum += bestcost
-            # step 3: move or not (in our case always move)
+
+                # print('Tour afterinsertion:', tmp)
+
+                # step 3: move or not (in our case always move)
             self.nodes = tmp
         return checksum
             

### 2020-09-30 implementations/numpy-flat-matrix/tspsolution.py
numpy implementations of lns benchmark

@@ -1,6 +1,9 @@
-import tspdata
-import numpy as np
+import sys
+import copy
 from itertools import chain
+import numpy as np
+
+import tspdata
 
 class TSPSolution:
     def __init__(self, data, permutation):
@@ -57,3 +60,33 @@ class TSPSolution:
                             tour[p+1:p+1+l] = t[:]
                         return True
         return False
+
+    def lns(self, niter=10):
+        d = self.data.d
+        checksum = 0
+        for iter in range(niter):
+            # step 0: copy solution
+            tmp = copy.copy(self.nodes)
+            unplanned = []
+            # step 1: destroy
+            where = 1
+            while where < len(tmp) - 1:
+                unplanned.append(tmp[where])
+                tmp = np.delete(tmp, where)
+                where += 1
+            # step 2: repair
+            while len(unplanned) > 0:
+                bestcost, bestnode, bestfro, bestto = sys.maxsize, -1, -1, -1
+                for (fro, k) in enumerate(unplanned):
+                    for ((pos, i), j) in zip(enumerate(tmp[:-1]), tmp[1:]):
+                        delta = d(i, k) + d(k, j) - d(i, j)
+                        if delta < bestcost:
+                            bestcost, bestfro, bestto = delta, fro, pos
+                # perform best found insertion
+                tmp = np.insert(tmp, bestto+1, bestnode)
+                unplanned = np.delete(unplanned, bestfro)
+                checksum += bestcost
+                # step 3: move or not (in our case always move)
+            self.nodes = tmp
+        return checksum
+    

### 2020-09-30 implementations/numpy/tspbenchmark.py

@@ -34,8 +34,8 @@ def loadfromfile(fname):
 def benchmarkone(solutions, benchmarkname):
     if benchmarkname == '2-opt':
         bench = lambda x: x.two_opt()
-    elif benchmarkname == 'LNS':
-        bench = lambda x: x.LNS()
+    elif benchmarkname == 'lns':
+        bench = lambda x: x.lns()
     elif benchmarkname == 'Or-opt':
         bench = lambda x: x.or_opt()
     else:

### 2020-09-30 implementations/numpy/tspsolution.py

@@ -1,6 +1,9 @@
-import tspdata
-import numpy as np
+import sys
+import copy
 from itertools import chain
+import numpy as np
+
+import tspdata
 
 class TSPSolution:
     def __init__(self, data, permutation):
@@ -57,6 +60,35 @@ class TSPSolution:
                             tour[p+1:p+1+l] = t[:]
                         return True
         return False
+
+    def lns(self, niter=10):
+        d = self.data.d
+        checksum = 0
+        for iter in range(niter):
+            # step 0: copy solution
+            tmp = copy.copy(self.nodes)
+            unplanned = []
+            # step 1: destroy
+            where = 1
+            while where < len(tmp) - 1:
+                unplanned.append(tmp[where])
+                tmp = np.delete(tmp, where)
+                where += 1
+            # step 2: repair
+            while len(unplanned) > 0:
+                bestcost, bestnode, bestfro, bestto = sys.maxsize, -1, -1, -1
+                for (fro, k) in enumerate(unplanned):
+                    for ((pos, i), j) in zip(enumerate(tmp[:-1]), tmp[1:]):
+                        delta = d[i][k] + d[k][j] - d[i][j]
+                        if delta < bestcost:
+                            bestcost, bestfro, bestto = delta, fro, pos
+                # perform best found insertion
+                tmp = np.insert(tmp, bestto+1, bestnode)
+                unplanned = np.delete(unplanned, bestfro)
+                checksum += bestcost
+                # step 3: move or not (in our case always move)
+            self.nodes = tmp
+        return checksum
     
     def dumpstring(self):
         return str(self.nodes)

### 2020-09-30 implementations/python/tspsolution.py

@@ -1,4 +1,4 @@
-import math
+import sys
 import copy
 from itertools import chain
 
@@ -63,9 +63,6 @@ class TSPSolution:
     def lns(self, niter=10):
         d = self.data.d
         checksum = 0
-
-        # print('Tour before:', self.nodes)
-        
         for iter in range(niter):
             # step 0: copy solution
             tmp = copy.copy(self.nodes)
@@ -76,41 +73,18 @@ class TSPSolution:
                 unplanned.append(tmp[where])
                 del tmp[where]
                 where += 1
-
-            # print('Tour after destruction:', tmp)
-        
             # step 2: repair
             while len(unplanned) > 0:
-                bestcost, bestnode, bestfro, bestto = math.inf, -1, -1, -1
+                bestcost, bestnode, bestfro, bestto = sys.maxsize, -1, -1, -1
                 for (fro, k) in enumerate(unplanned):
                     for ((pos, i), j) in zip(enumerate(tmp[:-1]), tmp[1:]):
-
-                        # print("Insertion of", k, "between", i, "and", j,
-                        #       "\tδ =", d[i][k] + d[k][j] - d[i][j])
-
-                        # print("\td[" + str(i) + "][" + str(k) + "] = "
-                        #         + str(d[i][k]))
-                        # print("\td[" + str(k) + "][" + str(j) + "] = "
-                        #         + str(d[k][j]))
-                        # print("\td[" + str(i) + "][" + str(j) + "] = "
-                        #         + str(d[i][j]))
-        
                         if d[i][k] + d[k][j] - d[i][j] < bestcost:
-
-                            # print('\tBEST')
-                            
                             bestcost = d[i][k] + d[k][j] - d[i][j]
                             bestnode, bestfro, bestto = k, fro, pos
                 # perform best found insertion
-
-                # print('Tour before insertion:', tmp)
-        
                 tmp.insert(bestto+1, bestnode)
                 del unplanned[bestfro]
                 checksum += bestcost
-
-                # print('Tour afterinsertion:', tmp)
-
                 # step 3: move or not (in our case always move)
             self.nodes = tmp
         return checksum

### 2020-09-30 implementations/python-flat-matrix/tspsolution.py
all cpython implementations of lns benchmark now done and improved

@@ -1,6 +1,8 @@
-import tspdata
+import sys
 import copy
 from itertools import chain
+
+import tspdata
 
 class TSPSolution:
     def __init__(self, data, permutation):
@@ -58,6 +60,36 @@ class TSPSolution:
                         return True
         return False
     
+    def lns(self, niter=10):
+        d = self.data.d
+        checksum = 0
+        for iter in range(niter):
+            # step 0: copy solution
+            tmp = copy.copy(self.nodes)
+            unplanned = []
+            # step 1: destroy
+            where = 1
+            while where < len(tmp) - 1:
+                unplanned.append(tmp[where])
+                del tmp[where]
+                where += 1
+            # step 2: repair
+            while len(unplanned) > 0:
+                bestcost, bestnode, bestfro, bestto = sys.maxsize, -1, -1, -1
+                for (fro, k) in enumerate(unplanned):
+                    for ((pos, i), j) in zip(enumerate(tmp[:-1]), tmp[1:]):
+                        delta = d(i, k) + d(k, j) - d(i, j)
+                        if delta < bestcost:
+                            bestcost = delta
+                            bestnode, bestfro, bestto = k, fro, pos
+                # perform best found insertion
+                tmp.insert(bestto+1, bestnode)
+                del unplanned[bestfro]
+                checksum += bestcost
+                # step 3: move or not (in our case always move)
+            self.nodes = tmp
+        return checksum
+
     def cost(self):
         total = 0
         for i, j in zip(self.nodes[:-1], self.nodes[1:]):

### 2020-09-30 implementations/python/tspsolution.py

@@ -78,8 +78,9 @@ class TSPSolution:
                 bestcost, bestnode, bestfro, bestto = sys.maxsize, -1, -1, -1
                 for (fro, k) in enumerate(unplanned):
                     for ((pos, i), j) in zip(enumerate(tmp[:-1]), tmp[1:]):
-                        if d[i][k] + d[k][j] - d[i][j] < bestcost:
-                            bestcost = d[i][k] + d[k][j] - d[i][j]
+                        delta = d[i][k] + d[k][j] - d[i][j]
+                        if delta < bestcost:
+                            bestcost = delta
                             bestnode, bestfro, bestto = k, fro, pos
                 # perform best found insertion
                 tmp.insert(bestto+1, bestnode)

### 2020-09-30 implementations/numba-flat-matrix/tspsolution.py
numba implementations of lns benchmark. Does not actually work due to the fact that numba does not yet implement np.insert

@@ -30,6 +30,11 @@ class TSPSolution:
         # actually speeds things up!
         tour, d, n = self.nodes, self.data.d, self.data.n
         return aux_oropt(tour, d, n)
+
+    def lns(self, niter=10):
+        return float('Inf')
+        tour, d, n = self.nodes, self.data.d, self.data.n
+        return aux_lns(tour, d, n, niter)    
     
 @jit(nopython=True)
 def aux_2opt(tour, d, n):
@@ -68,3 +73,34 @@ def aux_oropt(tour, d, n):
                             tour[p+1:p+1+l] = t[:]
                         return True
     return False
+
+# currently not working: np.insert not supported by numba...
+@jit(nopython=True)
+def aux_lns(tour, d, n, niter):
+    checksum = 0
+    for iter in range(niter):
+        # step 0: copy solution
+        tmp = tour.copy()
+        unplanned = []
+        # step 1: destroy
+        where = 1
+        while where < len(tmp) - 1:
+            unplanned.append(tmp[where])
+            tmp = np.delete(tmp, where)
+            where += 1
+        # step 2: repair
+        while len(unplanned) > 0:
+            bestcost, bestnode, bestfro, bestto = sys.maxsize, -1, -1, -1
+            for (fro, k) in enumerate(unplanned):
+                for ((pos, i), j) in zip(enumerate(tmp[:-1]), tmp[1:]):
+                    delta = dist(d, n, i, k) + dist(d, n, k, j) \
+                        - dist(d, n, i, j)
+                    if delta < bestcost:
+                        bestcost, bestfro, bestto = delta, fro, pos
+            # perform best found insertion
+            tmp = np.insert(tmp, bestto+1, bestnode)
+            del unplanned[bestfro]
+            checksum += bestcost
+            # step 3: move or not (in our case always move)
+        tour = tmp
+    return checksum

### 2020-09-30 implementations/numba/tspsolution.py

@@ -1,3 +1,4 @@
+import sys
 import itertools as it
 import numpy as np
 from numba import jit
@@ -30,6 +31,11 @@ class TSPSolution:
         # actually speeds things up!
         tour, d = self.nodes, self.data.d
         return aux_oropt(tour, d)
+
+    def lns(self, niter=10):
+        return float('Inf')
+        tour, d = self.nodes, self.data.d
+        return aux_lns(tour, d, niter)    
     
 @jit(nopython=True)
 def aux_2opt(tour, d):
@@ -68,3 +74,33 @@ def aux_oropt(tour, d):
                             tour[p+1:p+1+l] = t[:]
                         return True
     return False
+
+# currently not working: np.insert not supported by numba...
+@jit(nopython=True)
+def aux_lns(tour, d, niter):
+    checksum = 0
+    for iter in range(niter):
+        # step 0: copy solution
+        tmp = tour.copy()
+        unplanned = []
+        # step 1: destroy
+        where = 1
+        while where < len(tmp) - 1:
+            unplanned.append(tmp[where])
+            tmp = np.delete(tmp, where)
+            where += 1
+        # step 2: repair
+        while len(unplanned) > 0:
+            bestcost, bestnode, bestfro, bestto = sys.maxsize, -1, -1, -1
+            for (fro, k) in enumerate(unplanned):
+                for ((pos, i), j) in zip(enumerate(tmp[:-1]), tmp[1:]):
+                    delta = d[i][k] + d[k][j] - d[i][j]
+                    if delta < bestcost:
+                        bestcost, bestfro, bestto = delta, fro, pos
+            # perform best found insertion
+            tmp = np.insert(tmp, bestto+1, bestnode)
+            del unplanned[bestfro]
+            checksum += bestcost
+            # step 3: move or not (in our case always move)
+        tour = tmp
+    return checksum

### 2020-09-30 implementations/numpy/tspsolution.py

@@ -84,7 +84,7 @@ class TSPSolution:
                             bestcost, bestfro, bestto = delta, fro, pos
                 # perform best found insertion
                 tmp = np.insert(tmp, bestto+1, bestnode)
-                unplanned = np.delete(unplanned, bestfro)
+                del unplanned[bestfro]
                 checksum += bestcost
                 # step 3: move or not (in our case always move)
             self.nodes = tmp

### 2020-09-30 implementations/c++/tspsolution.h
rust implementation of lns

@@ -121,7 +121,6 @@ public:
 		T bestcost = INT_MAX;
 		for (unsigned int k=0; k < unplanned.size(); k++) {
 		    for (unsigned to=0; to < tmp.size() -1; to++) {
-			
 			T delta = data_->d(tmp[to], unplanned[k]) +
 			    data_->d(unplanned[k], tmp[to+1]) -
 			    data_->d(tmp[to], tmp[to+1]);

### 2020-09-30 implementations/rust/src/main.rs

@@ -100,7 +100,46 @@ impl TSPSolution {
         }
         return false;
     }
-    
+
+    fn lns(&mut self, data: &TSPData, niter: i32) -> i32 {
+        let mut checksum: i32 = 0;
+        for _iter in 1..(niter+1) {
+            // step 0: copy incumbent
+            let mut tmp = self.nodes.clone();
+            let mut unplanned = Vec::new();
+            // step 1: destroy
+            let mut t = 1;
+            while t < tmp.len() -1 {
+                unplanned.push(tmp[t]);
+                tmp.remove(t);
+                t += 1;
+            }
+            // step 2: repair
+            while unplanned.len() > 0 {
+                let mut bestfrom: usize = 0;
+                let mut bestto: usize = 0;
+                let mut bestcost: i32 = i32::MAX;
+                for (fro, k) in unplanned.iter().enumerate() {
+                    for (to, ij) in tmp.windows(2).enumerate() {
+                        let (i, j) = (ij[0], ij[1]);
+                        let delta = data.d(i, *k) + data.d(*k, j) - data.d(i, j);
+                        if delta < bestcost {
+                            bestcost = delta;
+                            bestfrom = fro;
+                            bestto = to;
+                        }
+                    }
+                }
+		// perform best found insertion
+		tmp.insert(bestto + 1, unplanned[bestfrom]);
+		unplanned.remove(bestfrom);
+		checksum += bestcost;
+            }
+            // step 3: set new incumbent
+            self.nodes = tmp;
+        }
+        return checksum;
+    }
 }
 
 fn read_data(fname: &str) -> (TSPData, Vec<TSPSolution>) {
@@ -149,6 +188,8 @@ fn benchmark_one(data: &TSPData, solutions: &mut Vec<TSPSolution>,
             cnt = sol.two_opt(data);
         } else if benchmarkname == "Or-opt" {
             cnt = sol.or_opt(data);
+        } else if benchmarkname == "lns" {
+            cnt = sol.lns(data, 10);
         } else {
             panic!("Unknown benchmark: {:?}", benchmarkname);
         }

### 2020-09-30 scripts/compare.R
added lns benchmark to scripts

@@ -18,7 +18,7 @@ if (! (reflang %in% languages)) {
     languages <- c(reflang, languages)
 }
 
-benchmarks <- c('2-opt', 'Or-opt')
+benchmarks <- c('2-opt', 'Or-opt', 'lns')
 
 ## first we glue all data together in one data frame
 fnames <- paste0(languages, '-runs.csv')

### 2020-09-30 scripts/runall.sh

@@ -13,7 +13,7 @@ if [[ "$#" -lt 1 ]]; then
     languages=$defaultlanguages
 fi
 
-benchmarks='2-opt Or-opt'
+benchmarks='2-opt Or-opt lns'
 
 # we do multi-threading!
 # Number of threads = environment variable THREADS or default value

### 2020-12-23 paper/main.tex
lns benchmark implementations: C++ (all except static), Java (all except static), Javascript
new note on data structures for Labels in Python

@@
 \section{Considered benchmarks and their format}
+\subsection{2-opt}
+\subsection{Or-opt}
+\subsection{Large neighbourhood search}
+
+Note: we use a vector and not a set to store unplanned requests,
+otherwise we take the risk of losing determinism.
+
+\subsection{Something DP-like}
 
@@
 
+\section{Other notes}
+On the ESPPRC benchmark, a label needs to store information about
+which customers have already been visited in the partial path it
+represents. In Python, preliminary testing reveals that using a set of
+integers for that purpose is twice as fast as using a vector of booleans.
+
 \bibliographystyle{plain}

### 2020-12-28 paper/main.tex
documentation of language differences in espprc and storing visited customers

@@
 represents. In Python, preliminary testing reveals that using a set of
-integers for that purpose is twice as fast as using a vector of booleans.
+integers for that purpose is twice as fast as using a vector of
+booleans. Similar preliminary testing in C++ reveals the opposite:
+using an array of booleans is faster. This is the behaviour we would
+expect in general, the observation on Python are the surprising ones
+here. There are two reasons for this expectation: (i) copying an array,
+represented in contiguous memory, can in general be performed faster
+than copying a set structure and (ii) lookup in an array is performed
+in $\cal{O}(1)$ time while lookup in a set is performed in $\cal{O}(log(n)$.
 

### 2020-12-28 implementations/julia-array/TSPBenchmark/src/TSPBenchmark.jl
bug fixes in Julia implementations

@@ -10,7 +10,6 @@ struct TSPData{ T<:Real }
     aux::Matrix{Float64}  # auxiliary graph used e.g. in column generation
 end
 
-include("tspdata.jl")
 include("espprc.jl")
 
 mutable struct TSPSolution

### 2020-12-28 implementations/julia-array/TSPBenchmark/src/espprc.jl

@@ -30,8 +30,8 @@ function dominates(l1::T, l2::T) where T <: Label
     if l1.cost > l2.cost || l1.length > l2.length
         return false
     end
-    for v ∈ l1.visits
-        if v ∉ l2.visits
+    for (v1, v2) ∈ zip(l1.visits, l2.visits)
+        if v1 > v2
             return false
         end
     end

### 2020-12-28 implementations/julia-square-matrix/TSPBenchmark/src/espprc.jl

@@ -30,8 +30,8 @@ function dominates(l1::T, l2::T) where T <: Label
     if l1.cost > l2.cost || l1.length > l2.length
         return false
     end
-    for v ∈ l1.visits
-        if v ∉ l2.visits
+    for (v1, v2) in zip(l1.visits, l2.visits)
+        if v1 > v2
             return false
         end
     end

### 2020-12-28 implementations/c++/espprc.h
in ESPPRC, various implementations: new label is now only added as successor to its predecessor if it is actually added and not immediately dominated

@@ -217,8 +217,8 @@ public:
 		    // at this point we know the extension is feasible
 		    Label<T> nl(label, succ, rc_, d_);
 		    auto ptr = make_shared<Label<T>>(nl);
-		    label->addsuccessor(ptr);
 		    bool added = Label<T>::update(labels[succ], nl);
+		    if (added) label->addsuccessor(ptr);
 		    if (added && (! inQ[succ]) && succ != 0) {
 			Q.push_back(succ);
 			inQ[succ] = true;

### 2020-12-28 implementations/c++98/espprc.h

@@ -220,8 +220,8 @@ public:
 		    }		    
 		    // at this point we know the extension is feasible
 		    Label<T> *nl = new Label<T>(label, succ, rc_, d_);
-		    label->addsuccessor(nl);
 		    bool added = Label<T>::update(labels[succ], nl);
+		    if (added) label->addsuccessor(nl);
 		    if (added && (! inQ[succ]) && succ != 0) {
 			Q.push_back(succ);
 			inQ[succ] = true;

### 2020-12-28 implementations/java/ESPPRC.java

@@ -188,8 +188,8 @@ public class ESPPRC {
 		    }
 		    // at this point we know the extension is feasible
 		    Label nl = new Label(label, succ, rc_, d_);
-		    label.addsuccessor(nl);
 		    boolean added = Label.update(labels.get(succ), nl);
+		    if (added) label.addsuccessor(nl);
 		    if (added && (! inQ[succ]) && succ != 0) {
 			Q.addLast(succ);
 			inQ[succ] = true;

### 2020-12-28 implementations/javascript/tspbenchmark.js

@@ -198,8 +198,8 @@ function ESPPRC(n, rc, d, nresources, resourcecapacity, maxlen) {
                     if (! rfeas) continue;
                     // at this point we know the extension is feasible
                     nl = label.extend(succ);
-                    label.addsuccessor(nl);
                     var added = update(labels[succ], nl);
+                    if (added) label.addsuccessor(nl);
                     if ( added && ( ! inQ[succ] && succ != 0 ) ) {
                         Q.push(succ);
                         inQ[succ] = true;

### 2020-12-28 implementations/python/espprc.py

@@ -43,8 +43,6 @@ class Label:
         # resource is consumed if i^th bit of vertex is 1
         nl.q = [ r + 1 if (vertex & (1 << i) > 0) else r
                  for i, r in enumerate(self.q) ]
-        #
-        self.successors.append(nl)
         return nl
 
     def __repr__(self):
@@ -126,6 +124,8 @@ class ESPPRC:
                     nl = label.extend(succ, self.rc, self.d)
                     # Let's update dominance
                     added = updatedominance(labels[succ], nl)
+                    if added:
+                        label.successors.append(nl)
                     if added and not (succ in Qset) and succ != 0:
                         Qset.add(succ)
                         Q.append(succ)

### 2020-12-28 paper/main.tex
explanation on Numba being not yet feature-complete

@@
 study. However it is worth noting that this can be a concern in general.
+\subsection{Numba}
+Numba is not yet feature-complete for Numpy. A direct consequence is that some
+benchmarks cannot be implemented for Numba. For example $np.insert()$ is not
+implemented, but this function is needed for the $LN$
+benchmark. Therefore the $LNS$ benchmark cannot be implemented with
+Numba. Still, Numba is significant enough to be benchmarked and its
+current performance are an indication of what to expect in years to come.
 

### 2021-01-14 paper/main.tex
style improvement
more description of implementation-specific quirks

@@
 \subsection{C++}
+Two implementations: c++14 vs c++98.
+
 \subsection{Python}
+Python vs Pypy vs Numpy vs Numba.
+
 \subsection{Java}
+
 \subsection{Julia}
+General performance tips: read the specific doc section! One unique
+official document outlines how to write efficient code.
+
 \subsection{Rust}
+
 \subsection{JavaScript}
@@
 
+\subsection{General cross-language considerations}
+Subjective considerations:
+\begin{itemize}
+\item Typically easier to program using Python, Julia or JavaScript.
+\item Typically harder to program using Rust than anything else but
+  might be due to lack of experience. However my experience in
+  JavaScript is much lower than in C++ and I still did everything
+  faster in JavaScript than in C++.
+\end{itemize}
+
 \section{Considered benchmarks and their format}
@@
 
-\subsection{Something DP-like}
+\subsection{Dynamic programming for ESPPRC}
+Motivation: column generation.
+
+Mention here espprc-index and how rust does not allow multiple
+references to other labels.
+
+\subsection{Maximum flow problem}
+Motivation: branch-and-cut.
 
@@
 
-\section{Language-specific notes}
+\section{Implementation-specific notes}
 \subsection{Python}
@@
 multiplication used for 0-indexed languages.
+
+Using two threads instead of 1 improves performance by about 10\%,
+although it is unclear what the second thread is used for.
+
 \subsection{JavaScript}
@@
 study. However it is worth noting that this can be a concern in general.
+
 \subsection{Numba}
 Numba is not yet feature-complete for Numpy. A direct consequence is that some
-benchmarks cannot be implemented for Numba. For example $np.insert()$ is not
-implemented, but this function is needed for the $LN$
-benchmark. Therefore the $LNS$ benchmark cannot be implemented with
-Numba. Still, Numba is significant enough to be benchmarked and its
-current performance are an indication of what to expect in years to come.
+benchmarks cannot be implemented for Numba. For example
+\lstinline{np.insert()} is not implemented, but this function is
+needed for the $LN$ benchmark. Therefore the $LNS$ benchmark cannot be
+implemented with Numba. Still, Numba is significant enough to be
+benchmarked and its current performance are an indication of what to
+expect in years to come.
+
+\subsection{Java}
+Mention inconsistencies between Java versions: code written for Java 8
+does not compile with Java 11. Because Pair is not in the API any more.
+
+\subsection{Rust}
+Issues:
+\begin{itemize}
+\item Changing from i32 to i64 is problematic somehow
+\item Not possible to have refs to both successors and predecessor in
+  Label, hence had to handle indices by hand in specific espprc implementation
+\end{itemize}
 
@@
 
+\section*{Acknowledgements}
+Sebastian for advice and help with Rust.
+
 \bibliographystyle{plain}

### 2021-01-14 implementations/c++/tspsolution.h
maxflow benchmar now divides all capacities by 1000. This allows to store checksum on 32 bit integers

@@ -198,7 +198,7 @@ public:
 	}
 	for (int i=0; i < n; i++) {
 	    for (int j=0; j < n; j++) {
-		C[i*n+j] = d[i*n+j] > t[j] ? (double) d[i*n+j] : 0.0;
+		C[i*n+j] = d[i*n+j] > t[j] ? d[i*n+j] / 1000.0 : 0.0;
 	    }
 	}
         // solve maxflow for each non-0 node as sink

### 2021-01-14 implementations/c++98/tspsolution.h

@@ -188,7 +188,7 @@ public:
 	}
 	for (int i=0; i < n; i++) {
 	    for (int j=0; j < n; j++) {
-		C[i*n+j] = d[i*n+j] > t[j] ? (double) d[i*n+j] : 0.0;
+		C[i*n+j] = d[i*n+j] > t[j] ? d[i*n+j] / 1000.0 : 0.0;
 	    }
 	}
         // solve maxflow for each non-0 node as sink

### 2021-01-14 implementations/java/TSPSolution.java

@@ -180,7 +180,7 @@ public class TSPSolution {
 	}
 	for (int i=0; i < n; i++) {
 	    for (int j=0; j < n; j++) {
-		C[i*n+j] = d[i*n+j] > t[j] ? (double) d[i*n+j] : 0.0;
+		C[i*n+j] = d[i*n+j] > t[j] ? d[i*n+j] / 1000.0 : 0.0;
 	    }
 	}
 	// solve maxflow for each non-0 sink

### 2021-01-14 implementations/javascript/tspbenchmark.js

@@ -173,7 +173,7 @@ function TSPSolution(data, sequence) {
         }
         for (var i=0; i < data.n; i++) {
             for (var j=0; j < data.n; j++) {
-                C[i*n+j] = d.d(i, j) > t[j] ? 1.0 * d.d(i, j) : 0.0;
+                C[i*n+j] = d.d(i, j) > t[j] ? d.d(i, j) / 1000.0 : 0.0;
             }
         }
         var checksum = 0.0;
@@ -181,7 +181,7 @@ function TSPSolution(data, sequence) {
             var mf = edmondskarp(C, F, n, 0, sink);
             checksum += mf;
         }
-        return Math.round(checksum);
+        return Math.trunc(checksum);
     }
 }
 

### 2021-01-14 implementations/julia-array/TSPBenchmark/src/tsp.jl

@@ -172,7 +172,7 @@ function maxflow(d::T, sol::ST) where T <: TSPData where ST <: TSPSolution
         t[j] = d.d[i, j]
     end
     for i ∈ 1:d.n, j ∈ 1:d.n
-        d.aux[i, j] = if (d.d[i, j] > t[j]) d.d[i,j] else zero(Float64) end
+        d.aux[i, j] = if (d.d[i, j] > t[j]) d.d[i,j]/1000 else zero(Float64) end
     end
     checksum = zero(Float64);
     for sink in 2:d.n

### 2021-01-14 implementations/julia-square-matrix/TSPBenchmark/src/tspsolution.jl

@@ -137,7 +137,8 @@ function maxflow(d::T, sol::ST) where T <: TSPData where ST <: TSPSolution
     end
     for i ∈ 1:d.n, j ∈ 1:d.n
         setaux!(d, i, j,
-                if (dist(d, i, j) > t[j]) dist(d, i, j) else zero(Float64) end)
+                if (dist(d, i, j) > t[j]) dist(d, i, j) / 1000 else
+                zero(Float64) end)
     end
     checksum = zero(Float64)
     for sink in 2:d.n

### 2021-01-14 implementations/python/tspsolution.py

@@ -131,7 +131,7 @@ class TSPSolution:
             t[j] = d[i][j]
         for j in range(n):
             for i in range(n):
-                cap[i][j] = float(d[i][j]) if d[i][j] > t[j] else 0.0
+                cap[i][j] = d[i][j] / 1000 if d[i][j] > t[j] else 0.0
         # then we solve it for each non-0 node as sink
         checksum = 0.0
         for sink in range(1, n):

### 2021-01-14 implementations/rust/src/tsp.rs

@@ -187,7 +187,7 @@ impl TSPSolution {
         for i in 0..data.n {
             for j in 0..data.n {
                 data.setaux(i, j, if data.d(i, j) > t[j]
-                            {f64::from(data.d(i, j))} else {0.0});
+                            {f64::from(data.d(i, j)) / 1000.0} else {0.0});
             }
         }
         // solve it for each non-0 node as sink

### 2021-01-14 implementations/numpy/espprcindex.py
numpy implementation of espprc-index

@@ -1,0 +1,125 @@
+import collections
+
+import numpy as np
+
+import espprc
+
+class LabelWithIndex(espprc.Label):
+    # construct an initial label corresponding to an empty path
+    def __init__(self):
+        # vertex where this partial path is arriving at
+        self.at = 0
+        # already visited nodes
+        self.visited = set()
+        # has this label already been extended?
+        self.ignore = False
+        # predecessor; should point to a Label
+        self.pred = -1
+        self.cost = 0
+        self.length = 0
+        # resource consumption
+        self.q = np.array([0 for x in self.resources])
+        # successors for later recursive deletion when dominated
+        self.successors = []
+
+    # Extend label to given vertex
+    def extend(self, rc, d, predindex, vertex):
+        nl = LabelWithIndex()
+        nl.at = vertex
+        nl.pred = predindex
+        nl.visited = self.visited.union((vertex,))
+        nl.cost = self.cost + rc[self.at,vertex]
+        nl.length = self.length + d[self.at,vertex]
+        # resource is consumed if i^th bit of vertex is 1
+        nl.q = np.array([ r + 1 if (vertex & (1 << i) > 0) else r
+                          for i, r in enumerate(self.q) ])
+        return nl
+
+    def __repr__(self):
+        return 'Label at node ' + str(self.at) \
+            + '\tcost = ' + str(self.cost) + '\tlength = ' + str(self.length) \
+            + '\tq = ' + str(self.q)
+    
+class LabelCollection:
+    def __init__(self):
+        self.labels = []
+        
+    def marksuccessors(self, lindex):
+        for s in self.labels[lindex].successors:
+            self.labels[s].ignore = True
+            self.marksuccessors(s)
+
+    # * compare newlabel with collection of labels removing dominated labels as
+    #   necessary
+    # * adds newlabel to collection labels if it is not dominated
+    # * returns true if newlabel was added
+    def updatedominance(self, labels, nlindex):
+        i = 0
+        while i < len(labels):
+            if self.labels[labels[i]].dominates(self.labels[nlindex]):
+                return False
+            elif self.labels[nlindex].dominates(self.labels[labels[i]]):
+                self.marksuccessors(labels[i])
+                if i < len(labels) - 1:
+                    labels[i] = labels[-1]
+                del labels[-1]
+            else:
+                i += 1
+        # at this point no label has dominated newlabel so we add it
+        labels.append(nlindex)
+        return True
+
+    def emptylabel(self):
+        self.labels.append(LabelWithIndex())
[…]

### 2021-01-14 implementations/numpy/tspbenchmark.py

@@ -42,6 +42,8 @@ def benchmarkone(solutions, benchmarkname):
         bench = lambda x: x.espprc(6, 1)
     elif benchmarkname == 'espprc-2':
         bench = lambda x: x.espprc(6, 2)
+    elif benchmarkname == 'espprc-index':
+        bench = lambda x: x.espprc(6, 1, True)
     else:
         bench = lambda x: None
     #

### 2021-01-14 implementations/numpy/tspsolution.py

@@ -5,6 +5,7 @@ import numpy as np
 
 import tspdata
 import espprc
+import espprcindex
 
 class TSPSolution:
     def __init__(self, data, permutation):
@@ -96,7 +97,7 @@ class TSPSolution:
     # the first bit being bit 0
     # For instance 6 consumes 1 unit of resource 1 and 1 unit of resource 2,
     # since 6 = 2^1 + 2^2
-    def espprc(self, nresources=6, resourcecapacity=1):
+    def espprc(self, nresources=6, resourcecapacity=1, index=False):
         n, tour, d, rc = self.data.n, self.nodes, self.data.d, self.data.aux
         # update reduced costs
         dual = np.array([ 0.0 for x in range(n) ])
@@ -108,9 +109,14 @@ class TSPSolution:
         # max len: sum of best assignments
         maxlen = sum([ min([d[i,j] for j in range(n) if i != j])
                        for i in range(n)])
-        e = espprc.ESPPRC(n, d, rc,
-                          self.nodes, nresources, resourcecapacity, maxlen)
-        return e.solve()
+        if not index:
+            e = espprc.ESPPRC(n, d, rc,
+                              self.nodes, nresources, resourcecapacity, maxlen)
+        else:
+            e = espprcindex.ESPPRCLC(n, d, rc,
+                                     self.nodes, nresources, resourcecapacity,
+                                     maxlen)
+        return int(e.solve())
     
     def dumpstring(self):
         return str(self.nodes)

### 2021-01-15 implementations/pypy/espprcindex.py
Pypy implementation of espprc-info and maxflow (symlinks to Python implementation)

@@ -1,0 +1,1 @@
+../python/espprcindex.py

### 2021-01-15 implementations/numpy/maxflow.py
numpy implementation of maxflow benchmark

@@ -1,0 +1,52 @@
+import sys
+import collections
+
+import numpy as np
+
+Inf = 100000
+
+# cap[i][j] is the capacity of arc (i, j)
+# flow[i][j]g is where we store the flow from i to j
+# n is the number of nodes in the complete graph
+# s is the source
+# t is the sink
+def edmondskarp(cap, flow, n, s, t):
+    totalflow = 0.0
+    moreflow = True
+    Q = collections.deque()
+    pred = np.array([ -1 for i in range(n) ])
+    
+    for i in range(n):
+        for j in range(n):
+            flow[i,j] = 0.0
+    # flow.fill(0.0)
+            
+    while moreflow:
+        # reset pred
+        for i in range(n):
+            pred[i] = -1
+        Q.append(s)
+        while len(Q) > 0:
+            cur = Q.popleft()
+            for j in range(n):
+                if j == cur:
+                    continue
+                if pred[j] == -1 and j != s and cap[cur,j] > flow[cur,j]:
+                    pred[j] = cur
+                    Q.append(j)
+        # did we find an augmenting path?
+        if pred[t] != -1:
+            df = Inf
+            i, j = pred[t], t
+            while i != -1:
+                if df > cap[i,j] - flow[i,j]:
+                    df = cap[i,j] - flow[i,j]
+                i, j = pred[i], i
+            i, j = pred[t], t
+            while i != -1:
+                flow[i,j] += df
+                i, j = pred[i], i
+            totalflow += df
+        else:
+            moreflow = False
+    return totalflow

### 2021-01-15 implementations/numpy/tspbenchmark.py

@@ -44,8 +44,11 @@ def benchmarkone(solutions, benchmarkname):
         bench = lambda x: x.espprc(6, 2)
     elif benchmarkname == 'espprc-index':
         bench = lambda x: x.espprc(6, 1, True)
+    elif benchmarkname == 'maxflow':
+        bench = lambda x: x.maxflow()
     else:
-        bench = lambda x: None
+        sys.stderr.write('Invalid benchmark: ' + benchmarkname + '\n')
+        sys.exit(22)
     #
     nimpr = 0
     totalcputime = 0.0

### 2021-01-15 implementations/numpy/tspdata.py

@@ -19,6 +19,7 @@ class TSPData:
         # place to have it.
         # dummy values to begin with...
         self.aux = numpy.array([ [ float(x) for x in row ] for row in d ])
+        self.aux2 = numpy.array([ [ float(x) for x in row ] for row in d ])
         
     def matrixstring(self):
         rows = [ ' '.join(str(x) for x in row) for row in self.d ]

### 2021-01-15 implementations/numpy/tspsolution.py

@@ -6,6 +6,7 @@ import numpy as np
 import tspdata
 import espprc
 import espprcindex
+import maxflow
 
 class TSPSolution:
     def __init__(self, data, permutation):
@@ -117,6 +118,26 @@ class TSPSolution:
                                      self.nodes, nresources, resourcecapacity,
                                      maxlen)
         return int(e.solve())
+
+    # max. flow
+    # arc capacity = arc cost divided by ten but only for those arcs that are
+    # above the average value of arcs with the same destination
+    def maxflow(self):
+        n, tour, d = self.data.n, self.nodes, self.data.d
+        cap, flow = self.data.aux, self.data.aux2
+        # first we build the capacity graph
+        t = np.array([ 0.0 for x in range(n) ])
+        for (i, j) in zip(tour[:-1], tour[1:]):
+            t[j] = d[i,j]
+        for j in range(n):
+            for i in range(n):
+                cap[i,j] = d[i,j] / 1000 if d[i,j] > t[j] else 0.0
+        # then we solve it for each non-0 node as sink
+        checksum = 0.0
+        for sink in range(1, n):
+            mf = maxflow.edmondskarp(cap, flow, n, 0, sink)
+            checksum += mf
+        return int(checksum)
     
     def dumpstring(self):
         return str(self.nodes)

### 2021-01-15 scripts/build_all.sh
Change in scripting: all "compile.sh" scripts have been renamed to "build.sh, which is more reflective of the true nature of what these scripts actually do, across all implementations (for instance in Julia they make sure that dependencies are installed) Scripts that call these scripts have been modified accordingly

@@ -10,13 +10,13 @@ impldir=`pwd`'/implementations'
 
 langs="c++ c++98 c++-hybrid-matrix c++-nested-matrix c++-static-arrays java java-nested-matrix javascript javascript-nested-matrix java-static-arrays julia-array julia-flat-matrix julia-square-matrix numba numba-flat-matrix numpy numpy-flat-matrix pypy pypy-nested-matrix python python-flat-matrix python-nested-matrix-function rust"
 
-echo "Compiling stuff"
+echo "Building stuff"
 for lang in $langs; do
-    (echo -e "Compiling \e[1m$lang\e[0m implementation"
+    (echo -e "Building \e[1m$lang\e[0m implementation"
      pushd $impldir'/'$lang > /dev/null
-     [[ -f compile.sh ]] && ./compile.sh
+     [[ -f build.sh ]] && ./build.sh
      popd > /dev/null
-     echo -e "Done compiling \e[1m$lang\e[0m "
+     echo -e "Done building \e[1m$lang\e[0m "
     ) &
     # parallelisation stolen from
     #https://unix.stackexchange.com/questions/103920/parallelize-a-bash-for-loop

### 2021-01-15 scripts/runall.sh

@@ -24,7 +24,7 @@ for benchmark in "${!benchmarks[@]}"; do
     for lang in ${benchmarks[$benchmark]}; do
 	(echo -e "Running \e[1m$lang\e[0m implementation of \e[1m$benchmark\e[0m benchmark"
 	 (pushd $impldir'/'$lang > /dev/null
-	  # [[ -f compile.sh ]] && ./compile.sh
+	  # [[ -f build.sh ]] && ./build.sh
 	  echo "language,version,benchmark,instance,n,nsolutions,checksum,time"
 	  ./run_benchmark.sh $instdir $benchmark
 	  popd > /dev/null

### 2021-01-15 implementations/c++-hybrid-matrix/build.sh
fixing some broken symlinks following previous update

@@ -1,1 +1,1 @@
-../c++/compile.sh
+../c++/build.sh

### 2021-01-15 implementations/java-static-arrays/build.sh

@@ -1,1 +1,1 @@
-../java/compile.sh
+../java/build.sh

### 2021-01-15 implementations/julia-flat-matrix/build.sh

@@ -1,1 +1,1 @@
-../julia-square-matrix/compile.sh
+../julia-square-matrix/build.sh

### 2021-01-15 implementations/numba/maxflow.py
numba impementation of maxflow benchmark + small improvements

@@ -1,7 +1,7 @@
 import sys
-import collections
 
 import numpy as np
+from numba import jit
 
 Inf = 100000
 
@@ -10,10 +10,11 @@ Inf = 100000
 # n is the number of nodes in the complete graph
 # s is the source
 # t is the sink
+@jit(nopython=True)
 def edmondskarp(cap, flow, n, s, t):
     totalflow = 0.0
     moreflow = True
-    Q = collections.deque()
+    Q = []
     pred = np.array([ -1 for i in range(n) ])
     for i in range(n):
         for j in range(n):
@@ -24,7 +25,7 @@ def edmondskarp(cap, flow, n, s, t):
             pred[i] = -1
         Q.append(s)
         while len(Q) > 0:
-            cur = Q.popleft()
+            cur = Q.pop(0)
             for j in range(n):
                 if j == cur:
                     continue

### 2021-01-15 implementations/numba/tspbenchmark.py

@@ -34,12 +34,17 @@ def loadfromfile(fname):
 def benchmarkone(solutions, benchmarkname):
     if benchmarkname == '2-opt':
         bench = lambda x: x.two_opt()
-    elif benchmarkname == 'lns':
-        bench = lambda x: x.lns()
+    # elif benchmarkname == 'lns':
+    #     bench = lambda x: x.lns()
     elif benchmarkname == 'Or-opt':
         bench = lambda x: x.or_opt()
+    # elif benchmarkname == 'espprc-index':
+    #     bench = lambda x: x.espprc_index(nresources=6, resourcecapacity=1)
+    elif benchmarkname == 'maxflow':
+        bench = lambda x: x.maxflow()
     else:
-        bench = lambda x: None
+        sys.stderr.write('Invalid benchmark: ' + benchmarkname + '\n')
+        sys.exit(22)
     #
     nimpr = 0
     totalcputime = 0.0

### 2021-01-15 implementations/numba/tspsolution.py

@@ -4,6 +4,7 @@ import numpy as np
 from numba import jit
 
 import tspdata
+import maxflow
 
 class TSPSolution:
     def __init__(self, data, permutation):
@@ -33,16 +34,23 @@ class TSPSolution:
         return aux_oropt(tour, d)
 
     def lns(self, niter=10):
+        sys.stderr.write('Not valid: numba implementation of lns\n')
+        sys.stderr.write('\tdeleting from array is not implemented in numba\n')
+        sys.exit(95)
         return float('Inf')
         tour, d = self.nodes, self.data.d
-        return aux_lns(tour, d, niter)    
+        return aux_lns(tour, d, niter)
+    
+    def maxflow(self):
+        return aux_maxflow(self.data.n, self.data.d, self.data.aux,
+                           self.data.aux2, self.nodes)
     
 @jit(nopython=True)
 def aux_2opt(tour, d):
     for p1 in range(len(tour) - 3):
         for p2 in range(p1+2, len(tour) - 1):
-            if d[tour[p1]][tour[p1+1]] + d[tour[p2]][tour[p2+1]] > \
-               d[tour[p1]][tour[p2]] + d[tour[p1+1]][tour[p2+1]]:
+            if d[tour[p1],tour[p1+1]] + d[tour[p2],tour[p2+1]] > \
+               d[tour[p1],tour[p2]] + d[tour[p1+1],tour[p2+1]]:
                 # improving 2-exchange found
                 for i in range((p2-p1+1) // 2):
                     tour[p1+1+i], tour[p2-i] = tour[p2-i], tour[p1+1+i]
@@ -55,9 +63,9 @@ def aux_oropt(tour, d):
         for l in range(1, 1 + min(3, len(tour)-1-i)):
             for tmp in (range(i-1), range(i+l, len(tour)-1)):
                 for p in tmp:
-                    delta = d[tour[i-1]][tour[i+l]] + d[tour[p]][tour[i]] + \
-                        d[tour[i+l-1]][tour[p+1]] - d[tour[p]][tour[p+1]]\
-                        - d[tour[i-1]][tour[i]] - d[tour[i+l-1]][tour[i+l]]
+                    delta = d[tour[i-1],tour[i+l]] + d[tour[p],tour[i]] + \
+                        d[tour[i+l-1],tour[p+1]] - d[tour[p],tour[p+1]]\
+                        - d[tour[i-1],tour[i]] - d[tour[i+l-1],tour[i+l]]
                     # perform improving move
                     if delta < 0:
[…]

### 2021-01-15 implementations/numpy/espprc.py

@@ -19,7 +19,7 @@ class Label:
         # resource consumption
         self.q = np.array([0 for x in self.resources])
         # successors for later recursive deletion when dominated
-        self.successors = []
+        self.successors = np.array([])
 
     # Returns True if self dominates other
     # pre-condition: self.at == other.at

### 2021-01-15 implementations/numpy/tspdata.py

@@ -20,31 +20,3 @@ class TSPData:
         # dummy values to begin with...
         self.aux = numpy.array([ [ float(x) for x in row ] for row in d ])
         self.aux2 = numpy.array([ [ float(x) for x in row ] for row in d ])
-        
-    def matrixstring(self):
-        rows = [ ' '.join(str(x) for x in row) for row in self.d ]
-        return '\n'.join(rows)
-    
-    def genrandom(self, n):
-        x = [ random.choice(range(mapsize)) for x in range(n) ]
-        y = [ random.choice(range(mapsize)) for x in range(n) ]
-        self.n = n
-        # Euclidean distances
-        self.d = [ [ int(math.hypot(x[i]-x[j], y[i]-y[j]) * 100)
-                     for j in range(n) ] for i in range(n) ]
-        # triangle inequality correction
-        for i in range(n):
-            for j in range(n):
-                if i == j: continue
-                for k in range(n):
-                    if i == k or j == k: continue
-                    if self.d[i][j] + self.d[j][k] < self.d[i][k]:
-                        self.d[i][k] = self.d[i][j] + self.d[j][k]
-
-
-    def genrandomcycle(self):
-        # solution generation: random permutations
-        # solutions always start and end with zero
-        points = [x for x in range(1, self.n)]
-        random.shuffle(points)
-        return [0] + points + [0]

### 2021-01-15 paper/main.tex
numba impementation of maxflow benchmark + small improvements
added explanation about why there is no Numba implementaion of ESPPRC benchmark

@@
 benchmarks cannot be implemented for Numba. For example
-\lstinline{np.insert()} is not implemented, but this function is
+\lstinline{numpy.insert()} is not implemented, but this function is
 needed for the $LN$ benchmark. Therefore the $LNS$ benchmark cannot be
-implemented with Numba. Still, Numba is significant enough to be
-benchmarked and its current performance are an indication of what to
-expect in years to come.
+implemented with Numba. Additionally, we were not able to implement
+the ESPPRC benchmark in a satisfying manner. In our experience, Numba
+error messages are sometimes unrelated to the issue causing them, or
+too obscure to make sense, keeping in mind the premises of this
+paper. It would certainly be possible to implement a dynamic
+programming algorithm for the ESPPRC benchmark using Numba's JIT
+compilationfeatures, however our conclusion is that it would not be
+possible to do so without expert knowledge of Numba.
+
+Still, Numba is significant enough to be
+benchmarked and its current performances are an indication of what to
+expect in years to come, as the project matures.
 
@@
 Mention inconsistencies between Java versions: code written for Java 8
-does not compile with Java 11. Because Pair is not in the API any more.
+does not compile with Java 11. Because Pair is not in the API any
+more.
 

### 2021-01-15 implementations/c++-hybrid-matrix/espprc.h
renamed c++ to c++14

@@ -1,1 +1,1 @@
-../c++/espprc.h
+../c++14/espprc.h

### 2021-01-15 implementations/c++-hybrid-matrix/main.cpp

@@ -1,1 +1,1 @@
-../c++/main.cpp
+../c++14/main.cpp

### 2021-01-15 implementations/c++-hybrid-matrix/run_benchmark.sh

@@ -1,1 +1,1 @@
-../c++/run_benchmark.sh
+../c++14/run_benchmark.sh

### 2021-01-15 implementations/c++-hybrid-matrix/tspsolution.h

@@ -1,1 +1,1 @@
-../c++/tspsolution.h
+../c++14/tspsolution.h

### 2021-01-15 implementations/c++-nested-matrix/build.sh

@@ -1,1 +1,1 @@
-../c++/build.sh
+../c++14/build.sh

### 2021-01-15 implementations/c++-static-arrays/tspdata.h

@@ -1,1 +1,1 @@
-../c++/tspdata.h
+../c++14/tspdata.h

### 2021-01-15 implementations/c++98/maxflow.h

@@ -1,1 +1,1 @@
-../c++/maxflow.h
+../c++14/maxflow.h

### 2021-01-15 scripts/runall.sh

@@ -10,13 +10,17 @@ instdir=`pwd`'/instances'
 # Number of threads = environment variable THREADS or default value
 [[ -z "$THREADS" ]] && THREADS=3
 
-all_langs="c++ c++98 c++-hybrid-matrix c++-nested-matrix c++-static-arrays java java-nested-matrix javascript javascript-nested-matrix java-static-arrays julia-array julia-flat-matrix julia-square-matrix numba numba-flat-matrix numpy numpy-flat-matrix pypy pypy-nested-matrix python python-flat-matrix python-nested-matrix-function rust"
-interesting_langs="c++ c++98 java javascript julia-flat-matrix julia-square-matrix numba pypy python rust"
-#keys='moo woof'
+all_langs="c++14 c++98 c++-hybrid-matrix c++-nested-matrix c++-static-arrays java java-nested-matrix javascript javascript-nested-matrix java-static-arrays julia-array julia-flat-matrix julia-square-matrix numba numba-flat-matrix numpy numpy-flat-matrix pypy pypy-nested-matrix python python-flat-matrix python-nested-matrix-function rust"
+lns_langs="c++14 c++98 java javascript julia_array julia-flat-matrix julia-square-matrix pypy python rust"
+espprc_langs="c++14 c++98 java javascript julia_array julia-flat-matrix julia-square-matrix pypy python"
+espprcindex_langs="c++14 c++98 java javascript julia_array julia-flat-matrix julia-square-matrix pypy python rust"
+maxflow_langs="c++14 c++98 java javascript julia_array julia-flat-matrix julia-square-matrix pypy python rust"
 declare -A benchmarks=( ["2-opt"]=$all_langs
 			["Or-opt"]=$all_langs
-			["lns"]=$interesting_langs
-			# ["espprc"]=$interesting_langs
+			["lns"]=$lns_langs
+			["espprc"]=$espprc_langs
+			["espprc-index"]=$espprcindex_langs
+			["maxflow"]=$maxflow_langs
 		      )
 
 echo "Running benchmarks"

### 2021-01-15 implementations/c++-hybrid-matrix/espprc-index.h
some files were apparently not added to the repo until now
bug fix in data object creation

@@ -1,0 +1,1 @@
+../c++14/espprc-index.h

### 2021-01-15 implementations/c++-hybrid-matrix/maxflow.h

@@ -1,0 +1,1 @@
+../c++14/maxflow.h

### 2021-01-15 implementations/c++-hybrid-matrix/tspdata.h

@@ -27,6 +27,7 @@ protected:
     T **dh_;
     // auxiliary graph for e.g. espprc
     double *aux_;
+    double *aux2_;
 
 public:
     TSPData(int n, T *d) {
@@ -35,6 +36,8 @@ public:
 	dh_ = new T*[n];
 	for (int i=0; i < n; i++)
 	    dh_[i] = d_ + n * i;
+        aux_ = new double[n*n];
+        aux2_ = new double[n*n];
     }
     // ~TSPData() { delete[] d_; }
     // load data from file
@@ -47,6 +50,7 @@ public:
     // used to store data for auxiliary graphs, typically needs to be
     // allocated once then overwritten many times
     double *aux() { return aux_; }
+    double *aux2() { return aux2_; }
 };
 
 #endif

### 2021-01-22 paper/main.tex
paper advancement

@@
 \label{sec:intro}
+
+Some languages are easier, but how much performance do we lose by
+using them?
+
 Purposes: determine which language we can tell our students to
@@
 (reading data, building initial solution) is not measured. Currently
-Julia also measures compilation time, needs to be fixed.
+Julia, Javascript, Pypy, Numba also measure compilation time.
 
-We restrict ourselves to variable-size vectors, since they are
-necessary in most algorithms for vehicle routing, for instance to
-insert or delete a vertex from a tour. This is the reason why we do
-not provide a C implementation. It is of course possible to
-implement variable-size vectors using fixed-size arrays and
-bookkeeping; however this stands directly against our stated purposes
-and philosophy. Moreover, implementing a bug-free, efficient version
-of variable-size vectors is unlikely to result in something better
-than existing libraries.
+In order to represent a tour in heuristics we restrict ourselves to
+variable-size vectors, since they are necessary in most algorithms for
+vehicle routing, for instance to insert or delete a vertex from a
+tour.
+% This is the reason why we do
+% not provide a C implementation. It is of course possible to
+% implement variable-size vectors using fixed-size arrays and
+% bookkeeping; however this stands directly against our stated purposes
+% and philosophy. Moreover, implementing a bug-free, efficient version
+% of variable-size vectors is unlikely to result in something better
+% than existing libraries.
+Inserting and deleting elements in a vector has linear worst-case time
+complexity. However, for heuristics it is still usually a good idea to
+use vectors rather than, say, linked lists, since neighborhood
+exploration typically requires to evaluate many moves then perform
+only one of these moves. The complexity of performing a move is thus
+secondary and unlikely to be a performance bottleneck. Additionally,
+performing certain moves can actually be more efficient using a
+vector than using a linked list, e.g. inverting a subsequence.
+
+\subsection{Limitations}
+Benchmarks are by definition limited in scope; any given benchmark
+measurement is only valid within a certain environment. In our case
+such environmental factors include the choice of compilers, compiler
+versions, operating system and CPU family. Therefore all results are
+to be taken as indicators but not absolute truths. The intent of this
+study is not to declare a language the winner because it is
+consistently faster than others by a few percents, but rather to
+assess what kind of performance losses can be expected by, for
+example, using Python instead of C++. OR to assess what kind of
+performance gains can be achieved by using a JIT compiler for Python.
+
+\section{The benchmarks}
+We consider five different benchmark algorithms that perform tasks
+commonly encountered in transport optimization. Together they capture
+key aspects of computationally intensive tasks in heuristics and
+exact methods for transport optimization. Additionally, we believe
+that they also capture such aspects for other fields of application
+of operations research, such as e.g. scheduling. For the sake of
+simplicity, we consider the symmetric travelling salesperson problem
+(TSP) as a base problem, i.e. input data are in the form of a
+symmetric distance matrix while a solution is a permutation. However
+some benchmark algorithms solve different problems than the TSP based on these
+TSP data, as described below.
+
+All five benchmarking algorithms described below are deterministic.
+Additionally, the input provided to them is also deterministic,
+i.e. each implementation receives the exact same inputs and performs
+the same operations from these inputs. For that purpose, we generate
+various instance files, each with a given size $n$, which represents
+the number of vertices in the graph, and a number $p$ of
+permutations. A permutation is a solution to the TSP. Each instance
+file contains one $n \times n$ distance matrix as well as $p$ randomly
+generated permutations, which are used as starting \emph{seeds} for
+the algorithms (e.g. as starting solution for 2-opt). Since the same
+instance files are given to 
+each implementation, all implementations perform the same operations
+and return the same result. In order to control result integrity we use a
+mechanism similar to \emph{checksum}. The checksum calculation for
+applying a certain algorithm using a given permutation as starting seed differs
+based on the algorithm and is explained separately for each algorithm
+below. The checksum calculation for an instance file is the sum of
+checksum values over all permutations in that instance file.
+
+Performances are measured per instance file, i.e. each time reported
+is for running an algorithm $p$ times, using one distance matrix with the $p$
+different seed permutations.
+
+\subsection{2-opt}
+One of the most commonly used heuristics in vehicle routing, 2-opt
+improves a tour by performing 2-exchanges. A 2-exchange consists in
+removing two edges from a tour and reconnecting the tour with two
+other edges. It is equivalent to inverting a sub-sequence of the
+tour, and can be performed in place using an array or vector solution
+representation.
+
+The checksum for a given permutation seed is calculated as the number of
+improvements found while applying first-improvement 2-opt, using this
+permutation as starting solution.
+
+\subsection{Or-opt}
+
+The checksum for a given permutation seed is calculated as the number of
+improvements found while applying first-improvement Or-opt, using this
+permutation as starting solution.
+
+\subsection{Large neighbourhood search}
+
+Note: we use a vector and not a set to store unplanned requests,
+otherwise we take the risk of losing determinism.
+
+The checksum for a given permutation seed is calculated as the total
+cost of insertions performed while applying 10 iterations of LNS using
+that permutation as starting solution.
+
+\subsection{Dynamic programming for ESPPRC}
+Motivation: column generation.
+
+Mention here espprc-index and how rust does not allow multiple
+references to other labels.
+
+The checksum for a given permutation seed is the cost of the shortest
+elementary path obtained when using that permutation seed to generate the
+ESPPRC instance, truncated to its integer part.
+
+\subsection{Maximum flow problem}
+Motivation: branch-and-cut.
+
+The checksum for a given permutation seed is the maximum flow value
+obtained when using that permutation seed to generate the MFP
+instance, truncated to its integer part.
 
 \section{Languages and implementations considered}
+We consider six different programming languages: C++, Python, Java,
+Julia, Rust and Javascript. Any rule of selection
+is arbitrary in nature; however, we indicate below, for each of these
+languages, the main reasons behind their selection. It is possible to
+add other languages to this list in the future.
+
+For each language, we may consider multiple \emph{implementations},
+which differ by which mechanisms or library they use. For
+instance with Python we can compare an implementation using native
+lists against an implementation using Numpy arrays. Implementations are
+detailed individually below for each language. The code for all
+implementations can be found at
+\url{https://github.com/fa-bien/routing-benchmark}.
+
 \subsection{C++}
@@
 
+Flat vs nested matrix.
+
+static arrays too.
+
 \subsection{Python}
 Python vs Pypy vs Numpy vs Numba.
+Flat vs nested matrix.
 
 \subsection{Java}
+Flat vs nested matrix.
+static arrays too.
 
@@
 General performance tips: read the specific doc section! One unique
-official document outlines how to write efficient code.
+official document outlines how to write efficient code. there is also
+an automated tool for that.
+
+Flat vs nested matrix.
 
@@
   computer with a modern web browser, including smartphones.
-\item Integrating JavaScript in a web interface is especially easy.
+\item Integrating JavaScript with a web interface is especially easy.
 \item JavaScript engines have received considerable attention from
@@
 command line. This means that there is in fact not even the need for a
-web browser. Node.js uses Google Chrome v8's JavaScript engine~\cite{nodejs}.
+web browser. Node.js currently uses Google Chrome v8's JavaScript
+engine~\cite{nodejs}.
 
-\subsection{General cross-language considerations}
-Subjective considerations:
-\begin{itemize}
-\item Typically easier to program using Python, Julia or JavaScript.
-\item Typically harder to program using Rust than anything else but
-  might be due to lack of experience. However my experience in
-  JavaScript is much lower than in C++ and I still did everything
-  faster in JavaScript than in C++.
-\end{itemize}
+Flat vs nested.
 
-\section{Considered benchmarks and their format}
-\subsection{2-opt}
-\subsection{Or-opt}
-\subsection{Large neighbourhood search}
-
-Note: we use a vector and not a set to store unplanned requests,
-otherwise we take the risk of losing determinism.
-
-\subsection{Dynamic programming for ESPPRC}
-Motivation: column generation.
-
-Mention here espprc-index and how rust does not allow multiple
-references to other labels.
-
-\subsection{Maximum flow problem}
-Motivation: branch-and-cut.
-
-\section{Considerations on how to implement a distance matrix}
+\subsection{Considerations on how to implement a distance matrix}
 Perhaps the most straightforward way to implement a distance matrix in
@@
 there should be no runtime penalty from using a function call. In
-general we expect a speedup from using a flat representation.
+general we expect a speedup from using a flat representation. We
+implemented some of the benchmarks with both flat and nested matrix
+representation in order to determine if there are significant
+performance differences.
 
-There are two cases where the above considerations on using a flat
-representation do not apply: Python does not allow inlining and Julia
-has native multi-dimensional arrays. This is discussed later.
+There are a few cases where the above considerations on using a flat
+representation do not apply: Python does not allow inlining, while
+Numpy and Julia provide multi-dimensional arrays. Nonetheless we also
+wrote a flat matrix version of these implementations, in order
+to elicit any difference in performance.
+
+\subsection{Other general cross-language considerations}
+Subjective considerations:
+\begin{itemize}
+\item Typically easier to program using Python, Julia or JavaScript.
+\item Typically harder to program using Rust than anything else but
+  might be due to lack of experience. However my experience in
+  JavaScript is much lower than in C++ and I still did everything
+  faster in JavaScript than in C++.
+\end{itemize}
 
@@
 remedy this.
+
+On the ESPPRC benchmark, a label needs to store information about
+which customers have already been visited in the partial path it
+represents. In Python, preliminary testing reveals that using a set of
+integers for that purpose is twice as fast as using a vector of
+booleans. Similar preliminary testing in C++ reveals the opposite:
+using an array of booleans is faster. This is the behaviour we would
+expect in general, the observation on Python are the surprising ones
+here. There are two reasons for this expectation: (i) copying an array,
+represented in contiguous memory, can in general be performed faster
+than copying a set structure and (ii) lookup in an array is performed
+in $\mathcal{O}(1)$ while lookup in a set is performed in
+$\mathcal{O}(log(n))$.
+
 \subsection{Julia}
@@
 
-\section{Other notes}
-On the ESPPRC benchmark, a label needs to store information about
-which customers have already been visited in the partial path it
-represents. In Python, preliminary testing reveals that using a set of
-integers for that purpose is twice as fast as using a vector of
-booleans. Similar preliminary testing in C++ reveals the opposite:
-using an array of booleans is faster. This is the behaviour we would
-expect in general, the observation on Python are the surprising ones
-here. There are two reasons for this expectation: (i) copying an array,
-represented in contiguous memory, can in general be performed faster
-than copying a set structure and (ii) lookup in an array is performed
-in $\cal{O}(1)$ time while lookup in a set is performed in $\cal{O}(log(n)$.
+\section{Experiments}
+All programs are run concurrently on an AMD Ryzen 5 3600 6-Core
+Processor at 3.6 GHz with 16 GB RAM running Linux. Twelve threads can
+run in parallel, however in our setting at most 5 threads are used at
+any given time.
+
+Mention here compilers and compiler versions used.
+
+\subsection{Impact of using a flat matrix}
+\subsection{Impact of using static arrays}
+\subsection{A comparison of Python implementations}
+\subsection{A comparison of C++ implementations}
+Also add a clang++ vs g++ comparison?
+\subsection{General cross-language comparison}
+\subsection{Cross-language comparison: fast languages}
+
+\section{Discussion}
+The question of what programming language to use when implementing
+optimization algorithms is one that, by essence, does not have a final
+answer.
 
 \section*{Acknowledgements}
-Sebastian for advice and help with Rust.
+The author is grateful to Sebastian Leitner for his valuable support
+in relation with the Rust implementation.
 

### 2021-01-25 implementations/julia-square-matrix/TSPBenchmark/src/maxflow.jl
those files should have been added in previous commits but were not :)

@@ -1,0 +1,46 @@
+function edmondskarp(d::MT, s::Int, t::Int) where MT <: TSPData
+    totalflow = zero(Float64)
+    moreflow = true
+    Q = Deque{Int}()
+    pred = [ -1 for i ∈ 1:d.n ]
+    for i ∈ 1:d.n, j ∈ 1:d.n
+        setaux2!(d, i, j, zero(Float64))
+    end
+    while moreflow
+        # reset predecessors
+        for i ∈ 1:d.n
+            pred[i] = -1
+        end
+        push!(Q, s)
+        while ! isempty(Q)
+            cur = popfirst!(Q)
+            for j in 1:d.n
+                j == cur && continue
+                if pred[j] == -1 && j ≠ s && getaux(d,cur,j) > getaux2(d,cur,j)
+                    pred[j] = cur
+                    push!(Q, j)
+                end
+            end
+        end
+        # did we find an augmenting path?
+        if pred[t] ≠ -1
+            df = typemax(Float64)
+            i, j = pred[t], t
+            while i ≠ -1
+                if df > getaux(d, i, j) - getaux2(d, i, j)
+                    df = getaux(d, i, j) - getaux2(d, i, j)
+                end
+                i, j = pred[i], i
+            end
+            i, j = pred[t], t
+            while i ≠ -1
+                addaux2!(d, i, j, df)
+                i, j = pred[i], i
+            end
+            totalflow += df
+        else
+            moreflow = false
+        end
+    end
+    totalflow
+end

### 2021-01-25 implementations/python/maxflow.py

@@ -1,0 +1,107 @@
+import sys
+import collections
+
+Inf = 100000
+zero = 0
+
+# cap[i][j] is the capacity of arc (i, j)
+# flow[i][j]g is where we store the flow from i to j
+# n is the number of nodes in the complete graph
+# s is the source
+# t is the sink
+def edmondskarp(cap, flow, n, s, t):
+    totalflow = 0.0
+    moreflow = True
+    Q = collections.deque()
+    pred = [ -1 for i in range(n) ]
+    for i in range(n):
+        for j in range(n):
+            flow[i][j] = 0.0
+    while moreflow:
+        # reset pred
+        for i in range(n):
+            pred[i] = -1
+        Q.append(s)
+        while len(Q) > 0:
+            cur = Q.popleft()
+            for j in range(n):
+                if j == cur:
+                    continue
+                if pred[j] == -1 and j != s and cap[cur][j] > flow[cur][j]:
+                    pred[j] = cur
+                    Q.append(j)
+        # did we find an augmenting path?
+        if pred[t] != -1:
+            df = Inf
+            i, j = pred[t], t
+            while i != -1:
+                if df > cap[i][j] - flow[i][j]:
+                    df = cap[i][j] - flow[i][j]
+                i, j = pred[i], i
+            i, j = pred[t], t
+            while i != -1:
+                flow[i][j] += df
+                i, j = pred[i], i
+            totalflow += df
+        else:
+            moreflow = False
+    return totalflow
+
+# taken and slightly adapted from wikipedia
+def relabel_to_front(C, F, n, source, sink):
+    for i in range(n):
+        for j in range(n):
+            F[i][j] = 0
+    # residual capacity from u to v is C[u][v] - F[u][v]
+
+    height = [0] * n  # height of node
+    excess = [0] * n  # flow into node minus flow from node
+    seen   = [0] * n  # neighbours seen since last relabel
+    # node "queue"
+    nodelist = [i for i in range(n) if i != source and i != sink]
+
+    def push(u, v):
+        send = min(excess[u], C[u][v] - F[u][v])
+        F[u][v] += send
+        F[v][u] -= send
+        excess[u] -= send
+        excess[v] += send
+
+    def relabel(u):
+        # Find smallest new height making a push possible,
+        # if such a push is possible at all.
+        min_height = Inf
+        for v in range(n):
+            if C[u][v] - F[u][v] > zero:
+                min_height = min(min_height, height[v])
+                height[u] = min_height + 1
+
+    def discharge(u):
[…]

### 2021-01-25 implementations/rust/src/maxflow.rs

@@ -1,0 +1,63 @@
+use std::collections::VecDeque;
+
+use crate::tsp;
+
+// d.aux contains capacities
+// d.aux2 is used to store flow
+// d.n is the number f nodes in the complete graph
+// s is the source
+// t is the sink
+pub fn edmondskarp(d: &mut tsp::TSPData, s: usize, t: usize) -> f64 {
+    let mut totalflow: f64 = 0.0;
+    let mut moreflow: bool = true;
+    let mut q: VecDeque<usize> = VecDeque::new();
+    // MAX means nil i.e. "no predecessor"
+    let mut pred: Vec<usize> = vec![usize::MAX; d.n];
+    for i in 0..d.n {
+        for j in 0..d.n {
+            d.setaux2(i, j, 0.0);
+        }
+    }
+    while moreflow {
+        // reset predecessors
+        for i in 0..d.n {
+            pred[i] = usize::MAX;
+        }
+        q.push_back(s);
+        while ! q.is_empty() {
+            let cur = q.pop_front().unwrap();
+            for j in 0..d.n {
+                if j == cur { continue }
+                if pred[j] == usize::MAX && j != s &&
+                    d.aux(cur, j) > d.aux2(cur, j) {
+                    pred[j] = cur;
+                    q.push_back(j);
+                }
+            }
+        }
+        // did we find an augmenting path?
+        if pred[t] != usize::MAX {
+            let mut df = f64::MAX;
+            let mut i = pred[t];
+            let mut j = t;
+            while i != usize::MAX {
+                if df > d.aux(i, j) - d.aux2(i, j) {
+                    df = d.aux(i, j) - d.aux2(i, j)
+                }
+                j = i;
+                i = pred[i];
+            }
+            i = pred[t];
+            j = t;
+            while i != usize::MAX {
+                d.setaux2(i, j, d.aux2(i, j) + df);
+                j = i;
+                i = pred[i];
+            }
+            totalflow += df;
+        } else {
+            moreflow = false;
+        }
+    }
+    return totalflow;
+}

### 2021-01-25 paper/main.tex
a bit more on the paper: description of Or-opt and LNS benchmarks

@@
 \subsection{2-opt}
+The 2-opt heuristic was first described by~\cite{2-opt}.
 One of the most commonly used heuristics in vehicle routing, 2-opt
@@
 tour, and can be performed in place using an array or vector solution
-representation.
+representation. If the distances are symmetric, then each move can be
+evaluated in constant time.
 
 The checksum for a given permutation seed is calculated as the number of
-improvements found while applying first-improvement 2-opt, using this
-permutation as starting solution.
+improvements found while applying first-improvement 2-opt until no
+improving move exists, using this permutation as starting solution.
 
 \subsection{Or-opt}
+Or-opt is a heuristic that was first introduced by~\cite(Or-opt}.
+It relies on the exploration of a neighborhood that is a subset of all
+3-exchanges. A 3-exchange consists in removing 3 edges from a tour and
+reconnecting the tour in a different way. Of all these moves, Or-opt
+only considers those that shift a sequence of 1, 2 or 3 vertices to a
+different position in the tour. Each move can be evaluated in constant time.
 
 The checksum for a given permutation seed is calculated as the number of
-improvements found while applying first-improvement Or-opt, using this
-permutation as starting solution.
+improvements found while applying first-improvement Or-opt until no
+improving move exists, using this permutation as starting solution.
 
 \subsection{Large neighbourhood search}
+Large neighborhood search (LNS) was originally introduced in
+conjunction with constraint programming to solve the vehicle routing
+problem with time windows (VRPTW)~\cite{lns}. Since it has been
+a very popular method for tackling a wide variety of vehicle routing
+problems, see e.g.~\cite{lnschapter}.
 
-Note: we use a vector and not a set to store unplanned requests,
-otherwise we take the risk of losing determinism.
+LNS iteratively (i) copies the \emph{incumbent} solution, (ii)
+destroys the copy by removing some vertices from it, (iii) repairs the
+partial solution by re-inserting previously removed customers in it
+and (iv) determines whether this newly produced solution becomes the
+new incumbent solution or not. The whole method relies, among other
+things, on randomness. This is an issue since we want each benchmark
+to be deterministic. To remedy this issue, we design deterministic
+schemes for both destroying and repairing stages. For each starting
+seed, we apply a certain number of LNS iterations. At each iteration,
+the vertices at even indices in the tour are removed. These removed
+vertices are stored in a vector in the order in which they were
+removed, i.e. from smallest to largest index. Then the cost of each
+possible insertion is calculated and the cheapest insertion is
+performed. Because removed vertices are stored in a vector, the
+algorithm's behaviour in case of same-cost insertions is
+deterministic.
 

### 2021-02-04 paper/main.tex
progress on paper

@@
 \usepackage{listings} % for inline code
+\usepackage{verbatim}
 
@@
 the number of vertices in the graph, and a number $p$ of
-permutations. A permutation is a solution to the TSP. Each instance
+permutations. A permutation is a solution to the TSP.
+Vertex coordinates are generated randomly in $[0, 100)$. Let $d_{ij}$
+be the Euclidean distance between $i$ and $j$, then the distance
+considered is $c_{ij} = \lfloor 100 d_{ij} \rfloor$. Since it is an
+integer number, it is sufficient to use integer number representation
+to compute the cost of a TSP solution; however certain benchmarks use
+floating-point number representation, as explained below.
+Each instance
 file contains one $n \times n$ distance matrix as well as $p$ randomly
 generated permutations, which are used as starting \emph{seeds} for
-the algorithms (e.g. as starting solution for 2-opt). Since the same
+the algorithms (e.g. as starting solution for 2-opt).
+
+Since the same
 instance files are given to 
@@
 \subsection{Or-opt}
-Or-opt is a heuristic that was first introduced by~\cite(Or-opt}.
+Or-opt is a heuristic that was first introduced by~\cite{Or-opt}.
 It relies on the exploration of a neighborhood that is a subset of all
@@
 
-\subsection{Large neighbourhood search}
+\subsection{LNS: Large neighbourhood search}
 Large neighborhood search (LNS) was originally introduced in
@@
 and (iv) determines whether this newly produced solution becomes the
-new incumbent solution or not. The whole method relies, among other
+new incumbent solution or not. The method relies, among other
 things, on randomness. This is an issue since we want each benchmark
@@
 
-\subsection{Dynamic programming for ESPPRC}
-Motivation: column generation.
+\subsection{ESPPRC: Dynamic programming for column generation}
+Many state-of-the-art exact methods for vehicle routing problems rely
+on column generation (see e.g.~\cite{Baldacci:2012survey}). In such
+methods, a significant amount of CPU effort is spent solving the
+\emph{pricing subproblem}, which differs depending on the problem at
+hand but is nonetheless often an elementary shortest path problem with
+resource constraints (ESPPRC). While it can be solved heuristically,
+in order to establish optimality it eventually needs to be solved
+exactly as well. For this benchmark we implement a dynamic
+programming algorithm similar to the one by~\cite{feillet:2004exact}
+to solve the ESPPRC, albeit without time windows. One difficulty in
+this problem stems from the existence of negative-cost cycles in the
+graph.
 
-Mention here espprc-index and how rust does not allow multiple
-references to other labels.
+In order to simulate an environment similar to the one encountered in
+column generation for VRPs, we derive an ESPPRC instance from each TSP
+solution by using an auxilliary distance matrix $c'$ from the
+existing distance matrix $c$, as follows:
+\begin{enumerate}
+\item For each vertex $j$, let $\pi_j$ be the cost of arc $(i, j)$,
+  where $i$ is the predecessor of $j$ in the TSP solution.
+\item For each arc $(i, j)$, set $c'_{ij}$ to $c_{ij} - d_j$.
+\end{enumerate}
+Additionally, we generate resource consumption as follows: we consider
+6 resources numbered from 0 to 5, and any vertex $i$ consumes resource
+$r$ iff the $r^{th}$ bit of $i$'s binary representation is 1
+(considering that the first bit is bit 0). That is, in C syntax, iff
+$i & (1 << r) > 0$. Capacity is set to 1 for each resource.
+The algorithm then computes the elementary shortest path with resource
+constraints on $c'$ from 0 to 0 (1 to 1 in Julia).
+
+For this benchmark, floating-point numbers are used to represent cost,
+even though the values are integer. This is to emulate the usual
+setting when solving such problems.
+
+Each label needs (i) a reference to its predecessor, so that
+paths can be reconstructed and (ii) a collection of references to
+its successors, in order to recursively propagate dominance and later
+ignore labels that are successors of an already proved-to-be-dominated
+label. Rust does not allow that, at least not in the intuitive way
+where a label is encapsulated in a \verb|struct|, including references
+to predecessor and successors. This is because Rust is focused on
+security, while this type of construct can lead to unsafe operations if
+not used properly. This means that this benchmark cannot be
+implemented with Rust, even though it is consistent and reliable with all
+other implementations. In order to remediate this issue, we also
+implement a second benchmark to implement the same algorithm, although
+in a less intuitive way: all labels are stored in a collection, and
+the indices in this collection are used in other collections to store all
+predecessors and successors. This second implementation is more
+tedious, but it may also be more efficient, so we also implement it in
+other languages. We call it \emph{espprc-index}.
 
@@
 
-\subsection{Maximum flow problem}
-Motivation: branch-and-cut.
+\subsection{Maxflow: Maximum flow problem}
+Another very popular framework for the exact solution of routing
+problems is branch and cut. \cite{padberg:532}
+
+Similar to the ESPPRC benchmark above and for the same reason, we use
+floating-point number representation for this benchmark.
 

### 2021-02-05 paper/main.tex
more paper progress

@@
 key aspects of computationally intensive tasks in heuristics and
-exact methods for transport optimization. Additionally, we believe
+exact methods for transport optimization, although they do not
+necessarily use the most efficient algorithm to solve the problem they
+are tackling. Additionally, we believe
 that they also capture such aspects for other fields of application
@@
 
-Since the same
-instance files are given to 
+Since the same instance files are given to 
 each implementation, all implementations perform the same operations
@@
 
-\subsection{ESPPRC: Dynamic programming for column generation}
+\subsection{Dynamic programming for column generation}
 Many state-of-the-art exact methods for vehicle routing problems rely
@@
 solution by using an auxilliary distance matrix $c'$ from the
-existing distance matrix $c$, as follows:
+distance matrix $c$, as follows:
 \begin{enumerate}
@@
 the indices in this collection are used in other collections to store all
-predecessors and successors. This second implementation is more
-tedious, but it may also be more efficient, so we also implement it in
-other languages. We call it \emph{espprc-index}.
+predecessors and successors. This second implementation represents a
+different benchmark, which we call \emph{espprc-index}, whereas the
+base benchmark is called \emph{espprc}. The version with indices is more
+tedious to implement, but it may also be more efficient, so we also
+implement it in other languages.
 
@@
 
-\subsection{Maxflow: Maximum flow problem}
-Another very popular framework for the exact solution of routing
-problems is branch and cut. \cite{padberg:532}
+\subsection{Maximum flow problem}
+Another popular framework for the exact solution of routing
+problems is \emph{branch and cut}. Branch and cut is typically applied
+to solve mixed-integer linear programs; it is akin to branch and bound
+but only considers a subset of all constraints explicitly. Other
+constraints are generated dynamically when they are found to be
+violated while exploring the search tree. These constraints are said
+to be \emph{separated}. One early success story of branch and cut is
+actually in application to the symmetric TSP~\citep{padberg:532}, by
+separating subtour elimination constraints. The separation procedure
+involves finding a minimum capacity cut, which itself involves solving
+a maximum flow problem. For this reason the maximum flow problem is
+important in vehicle routing, and we dedicate one benchmark to solving
+it. The \emph{maxflow} benchmark implements the Edmonds-Karp
+algorithm~\cite{edmonds-karp}, which is a specific version of the
+Ford-Fulkerson~\cite{ford-fulkerson} augmenting path method. The
+specificity resides in how the augmenting paths are generated.
 
-Similar to the ESPPRC benchmark above and for the same reason, we use
-floating-point number representation for this benchmark.
+We derive a capacity graph $C$ from the distance matrix $c$ and a
+given TSP solution as follows: 
+\begin{enumerate}
+\item For each vertex $j$, let $t_j$ be the cost of arc $(i, j)$,
+  where $i$ is the predecessor of $j$ in the TSP solution.
+\item For each arc $(i, j)$, set $C_{ij}$ to $c_{ij}/1000$ if $c_{ij}
+  > t_j$, otherwise set it to 0.
+\end{enumerate}
 
-The checksum for a given permutation seed is the maximum flow value
-obtained when using that permutation seed to generate the MFP
-instance, truncated to its integer part.
+For any starting seed, the capacity graph is generated then the
+Edmonds-Karp algorithm is applied with 0 as \emph{source} and every other
+node as \emph{sink}. The checksum for that permutation seed is the sum
+of all maximum flow values thus obtained, truncated to its integer part.
+
+The reason why capacity values are distance values divided by 1000 is
+that otherwise, checksum values can exceed the largest possible
+integer representation with 32 bits, which can be a problem with some
+languages. This means that floating-point numbers are used, which is
+representative of the use case of branch and cut for vehicle routing problems.
 

### 2021-02-11 paper/main.tex
progrss on the paper

@@
 \subsection{Dynamic programming for column generation}
+\label{sec:espprc}
 Many state-of-the-art exact methods for vehicle routing problems rely
@@
 (considering that the first bit is bit 0). That is, in C syntax, iff
-$i & (1 << r) > 0$. Capacity is set to 1 for each resource.
+\verb|i & (1 << r) > 0|. Capacity is set to 1 for each resource.
 The algorithm then computes the elementary shortest path with resource
@@
 \subsection{Other general cross-language considerations}
-Subjective considerations:
+Here we discuss considerations that are subjective in nature, as they
+relate to our experience implementing all benchmarks in all
+languages. We need to take into account our experience using various
+languages prior to the beginning of this work:
 \begin{itemize}
-\item Typically easier to program using Python, Julia or JavaScript.
-\item Typically harder to program using Rust than anything else but
-  might be due to lack of experience. However my experience in
-  JavaScript is much lower than in C++ and I still did everything
-  faster in JavaScript than in C++.
+\item C++: numerous projects over 20 years.
+\item Python: numerous projects over 15 years; no experience of Numpy
+  or Numba.
+\item Java: PhD work 15 years ago, one minor project 5 years ago.
+\item Julia: one ongoing research project started 3 years ago.
+\item Javascript: a few hobby projects 5-7 years ago.
+\item Rust: no experience at all.
 \end{itemize}
+To add to the subjectivity, each benchmark was first implemented in
+Python then in other languages, with Rust typically being the last one
+to be implemented. Therefore the languages implemented last were
+supposedly easier, due to increased familiarity with the
+algorithm. For all reasons mentioned above, measuring the
+time spent implementing each benchmark in each language would be
+meaningless, therefore we limit ourselves to subjective opinion here.
+Keeping this in mind, we see a subset of perceived
+\emph{easy} languages, as in ``easy to program in and get things running
+with the expected outcome'', which consists of Python, Julia and
+JavaScript. These 
+easy languages typically allow to program the same algorithms faster
+than the other ones with the same results, performance considerations
+aside.
+On the other hand, we found the process of getting benchmarks to work
+with Rust to be much more tedious than with any other language,
+although this is certainly due to lack of experience. Nonetheless, the
+JavaScript implementation was typically easier than the C++ or Java ones
+despite having more experience with C++ and Java, so
+experience does not explain everything.
+Ultimately, there is little doubt that any experienced programmer in
+any of these languages would have no trouble implementing any of these
+benchmarks. Nonetheless, this programmer with experience in mostly C++
+and Python implemented every benchmark faster in Python, JavaScript or Julia
+than in C++.
+
+We also want to underline that no matter the language, it is possible
+to write both efficient and inefficient code. Familiarity and
+experience are beneficial; beginners in a language are more likely to
+write inefficient code. This being said, not every language is equal
+when it comes to documenting how to write efficient code. We note that
+the official Julia documentation has a whole section dedicated to that
+topic~\cite{julia-performance}. There also exists an automated tool to
+detect inefficient code patterns, based on the official
+documentation~\cite{julia-traceur}.
 
@@
 benchmarks cannot be implemented for Numba. For example
-\lstinline{numpy.insert()} is not implemented, but this function is
-needed for the $LN$ benchmark. Therefore the $LNS$ benchmark cannot be
+\verb|numpy.insert()| is not implemented, but this function is
+needed for the $LNS$ benchmark, therefore the $LNS$ benchmark cannot be
 implemented with Numba. Additionally, we were not able to implement
@@
 programming algorithm for the ESPPRC benchmark using Numba's JIT
-compilationfeatures, however our conclusion is that it would not be
-possible to do so without expert knowledge of Numba.
+compilation features, however our conclusion is that it would not be
+possible to do so without advanced knowledge of Numba.
 
@@
 \subsection{Java}
-Mention inconsistencies between Java versions: code written for Java 8
-does not compile with Java 11. Because Pair is not in the API any
-more.
+The ecosystem of Java implementations and versions can be hard to
+navigate. There exist multiple compilers, multiple virtual
+machines, and there is a new Java version every 6 months, sometimes
+breaking backwards compatibility. The official reference
+implementation is OpenJDK since version 7~\cite{openjdk-reference}.
+We use OpenJDK version 11, which is the current long-term support version as
+of writing this article.
 
 \subsection{Rust}
-Issues:
-\begin{itemize}
-\item Changing from i32 to i64 is problematic somehow
-\item Not possible to have refs to both successors and predecessor in
-  Label, hence had to handle indices by hand in specific espprc implementation
-\end{itemize}
+As mentioned in Section~\ref{sec:espprc}, Rust does not allow a given
+\verb|struct| to have multiple references to the same \verb|struct|
+type. For this reason, we needed to develop a different implementation
+of the espprc benchmark. In general Rust puts restrictions on what
+constitutes valid code, for the sake of safety. This is a valuable
+feature in systems programming, which is the main application that Rust was
+designed for. This also means that writing code that compiles and runs
+is actually a more tedious task, but that once the code compiles,
+there is also a higher chance that it will run according to
+expectations, like with the Ada programming language for instance.
+
+In the context of this study, these benefits did not appear relevant
+however: each benchmark was programmed in all other languages as well,
+providing the same results as with the Rust implementation, while no
+evidence of memory leak was observed. As mentioned above, The Rust
+implementation was typically way more time-consuming than other
+implementations, although one reason for that is lack of
+experience. Nonetheless, in the context of this study, the generally
+assumed benefits of using Rust are not apparent, with the notable
+exception of good runtime performance.
 
@@
 run in parallel, however in our setting at most 5 threads are used at
-any given time.
+any given time. For every benchmark and every instance, CPU times are
+reported as a ratio of a \emph{reference time} for that benchmark and
+instance. In practice, for each benchmark and instance, the reference
+time is the time required by the c++98 implementation for that
+benchmark and that instance. Therefore the c++98 implementation always
+has a CPU ratio of 1.
 
@@
 \subsection{Cross-language comparison: fast languages}
+
+\subsection{Cross-platform comparison}
+We now look at the same benchmarks, using the same languages, but running
+in a different environment. We compare runs on an Intel i5-8265U
+4-core CPU at 1.6 GHz and on the previously mentioned AMD CPU. The
+goal is not to observe which CPU provides the best performance, rather
+to determine whether the experimental observations made with one CPU
+are valid with the other. For this reason, we stick to reporting CPU
+times as a ratio to the CPU time required by the c++98 implementation.
+
+Perhaps also run on WU cluster and observe differences.
 

### 2021-02-15 paper/main.tex
more paper content

@@
 implementation in the world, but a reasonably good one, i.e. one we
-can expect from an average student. Similar algorithm in various languages.
+can expect from an average student. Similar algorithm in various
+languages. Use as few external libraries as possible, but use them if
+needed.
 
-What we measure: run time of the algorithm only. Everything else
-(reading data, building initial solution) is not measured. Currently
-Julia, Javascript, Pypy, Numba also measure compilation time.
+\subsection{What we measure}
+In all proposed benchmarks, we measure the CPU clock time of the
+benchmarked algorithm. Every other task, such as reading input data or
+constructing an initial solution for local search benchmarks, is not
+measured. However if an auxilliary graph needs to be constructed every
+time the algorithm is run, then we also measure the construction of
+said auxilliary graph. This is the case for benchmarks \emph{maxflow}
+and \emph{espprc}.
 
-In order to represent a tour in heuristics we restrict ourselves to
-variable-size vectors, since they are necessary in most algorithms for
+In general, compilation time is not measured. However, some
+implementations do just-in-time (JIT) compilation, for example Julia,
+JavaScript, Pypy, Numba and Java. In such cases, this JIT compilation time is
+included in the running time of the algorithm. In general we measure
+clock time, i.e. the CPU budget used by the algorithm. In the case of
+JavaScript we measure wall time as we are not aware of any
+way to measure clock time in JavaScript.
+
+In order to represent a tour in heuristics we use variable-size
+vectors, since they are necessary in most algorithms for 
 vehicle routing, for instance to insert or delete a vertex from a
@@
 assess what kind of performance losses can be expected by, for
-example, using Python instead of C++. OR to assess what kind of
+example, using Python instead of C++. Another example is to assess what kind of
 performance gains can be achieved by using a JIT compiler for Python.

### 2021-02-17 paper/main.tex
paper progress

@@
 
-Philosophy: not necessarily the best
-implementation in the world, but a reasonably good one, i.e. one we
-can expect from an average student. Similar algorithm in various
-languages. Use as few external libraries as possible, but use them if
-needed.
+\subsection{Scope of this study}
+We want to compare the reasonably efficient implementation of various
+algorithms used in combinatorial optimization. In particular, we look
+at five algorithms that are commonly used in vehicle routing
+optimization. However these algorithms have features that are also
+present in other domains of optimization, such as modifying
+subsequences, inserting and removing elements in a sequence, or other
+operations on graphs.
+
+Our goal is not to provide the most efficient implementation of each
+algorithm for a set of languages, or even to use the most efficient
+algorithm to solve the given problems, but rather to use these
+algorithms as benchmarks, each capturing certain features currently
+encountered in optimization. We put ourselves in the situation of an
+average student having to implement these algorithms, and compare the
+performance of different programming languages in that context. We
+also assess the impact on performance of some implementation
+decisions, e.g. using a flat matrix representation.
 
@@
 In order to represent a tour in heuristics we use variable-size
-vectors, since they are necessary in most algorithms for 
+vectors. Dynamic size containers are necessary in most algorithms for 
 vehicle routing, for instance to insert or delete a vertex from a
@@
 performing certain moves can actually be more efficient using a
-vector than using a linked list, e.g. inverting a subsequence.
+vector than using a linked list, e.g. inverting a subsequence in the
+context of 2-opt.
 
@@
 Julia, Rust and Javascript. Any rule of selection
-is arbitrary in nature; however, we indicate below, for each of these
-languages, the main reasons behind their selection. It is possible to
-add other languages to this list in the future.
+is arbitrary in nature; however we indicate below, for each of these
+languages, our perceived advantages of these languages, which justify
+their selection. It is possible to add other languages to this list in
+the future, and keep the benchmark collection alive in an online form.
+For a given language several implementations are possible, e.g. one
+with nested distance matrices and one with flat distance matrices. As
+a convention, implementations are named in lower-case letters, for
+example \emph{python-flat-matrix}. We also use the simplest names for
+the best implementations, i.e. \emph{python} uses a nested distance
+matrix because it performs better than \emph{python-flat-matrix}, but
+\emph{pypy} uses a flat matrix because it performs better than
+\emph{pypy-nested-matrix}. All implementations are summarised in
+Appendix~\ref{app:impl}.
 
@@
 
+It is difficult to properly measure the popularity of a programming
+language. There exist rankings, for example based on online
+searches~\cite{TIOBE} or on GitHub contributions~\cite{octoverse}. All
+these approaches have flaws. Basing an index on online searches is
+subject to all the biases and exploits that target online
+search. Measuring GitHub contributions is not clear, as
+there is no trivial way to decide what should count among
+repositories, number of lines written, number of commits,
+etc. Ultimately, language popularity for its own sake does not need to
+be a crucial factor in the choice of a programming language, especially for
+academic research. These aspects seem more important to us:
+\begin{itemize}
+\item The programs written in a given programming language should be
+  easily run on any computer. For that reason, we only look at
+  languages that are free and work under multiple operating
+  systems. The ability to run under Linux is especially important,
+  since it is typically the operating system of high-performance
+  computing (HPC) clusters~\cite{top500}.
+\item The language should be well documented, but also have online
+  resources for support. This in fact also depends on popularity.
+\end{itemize}
+
 \subsection{C++}
-Two implementations: c++14 vs c++98.
+C++ is one of the most popular programming languages in the World, and
+has been so for decades. Among its many features there is good
+runtime performance. In fact the performance is good enough that whole
+operating systems can be written in C++, for example Microsoft
+Windows. In fact C++ has so many features that it is often possible to
+implement the same algorithm using different subsets of the language,
+and a frequent criticism is that people and teams evolve into using
+only their own subset of the language.
 
-Flat vs nested matrix.
+We develop two C++ implementations of every benchmark, using the C++98
+and C++14 standards, which we call \emph{c++98} and \emph{c++14}. they
+both use a flat distance matrix. With C++98 we use references to
+objects, while with C++14 we use smart pointers such as
+\verb|shared_ptr|. Comparing these two implementations will allow us
+to evaluate the cost of smart pointers.
 
-static arrays too.
+For each benchmark the reference CPU time is provided by the c++14
+implementation. We also develop a variant of c++14 using a nested
+distance matrix, and another using static arrays.
 
 \subsection{Python}
-Python vs Pypy vs Numpy vs Numba.
-Flat vs nested matrix.
+Python is another of the most popular programming languages in the
+World, especially for tasks of machine learning in recent years. The
+base Python implementation is called \emph{python} and uses a nested
+distance matrix. There is also an implementation using a flat distance
+matrix, called \emph{python-flat-matrix}, and one using a nested
+distance matrix wrapped in function calls, called
+\emph{python-nested-matrix-function}.
+
+One issue with Python is that much of the code is not compiled, rather
+directly interpreted, which can lead to poor performance. One way to
+address that issue is to rely on libraries that are implemented in C,
+however that is not always possible. We consider alternative
+implementations in Python, all bringing different solutions to this
+issue.
+
+The simplest way to speed up the run time of a Python program is
+to run this program with Pypy~\cite{pypy} instead of the
+standard CPython interpreter. Pypy uses JIT compilation to speed
+things up. There is no need to modify the Python program, which is a
+strong advantage compared to library-based solutions or
+annotation-based solutions. Pypy's website claims that it is on
+average 4.2 times faster than CPython~\cite{pypy}, however this is
+just an average and actual numbers may vary a lot depending on the
+benchmark. We use both the python and python-flat-matrix
+implementations with Pypy, resulting in \emph{pypy-nested-matrix} and
+\emph{pypy} implementations, respectively.
+
+We also implement the benchmarks using Numpy arrays instead
+of standard Python lists. Numpy is one of the most popular external
+libraries for Python; it is coded in C for runtime
+performance~\cite{numpy}. Numpy performs especially well when using
+\emph{array operations}, for instance matrix multiplication. However
+the benchmarks that we consider do not take advantage of Numpy
+functions, only of its array data structure. Since Numpy provides
+multi-dimensional arrays, we use a two-dimensional array to represent
+the distancematrix. Additionally, we consider a flat matrix in
+implementation \emph{numpy-flat-matrix}.
+
+Numpy itself does not provide any kind of JIT compilation, but another
+Python package, Numba, performs JIT compilation when using Numpy
+arrays. Numba requires code annotation, therefore we annotate the
+Numpy implementations to take advantage of them with Numba.
 
 \subsection{Java}
-Flat vs nested matrix.
-static arrays too.
+Java is a very popular programming language of the past two
+decades. Java code is compiled to \emph{bytecode}, which is then run
+by a virtual machine (VM). The bytecode compiled on one computer can
+be run by a VM on another computer. Nonetheless, JIT compilation to
+native code happens dynamically at run time. The syntax is based on
+C/C++ syntax. The base Java implementation, \emph{java}, uses a flat
+distance matrix and a \verb|ArrayList<Integer>| object to represent a
+tour. We also implement a version with nested distance matrix,
+\emph{java-nested-matrix}, as well as a version with static arrays to
+represent a tour, \emph{java-static-arrays}.
 
 \subsection{Julia}
-General performance tips: read the specific doc section! One unique
-official document outlines how to write efficient code. there is also
-an automated tool for that.
+Julia is a relatively recent programming language. Mention dates for
+first release (0.4?) as well as 1.0.
 
-Flat vs nested matrix.
+Flat vs square matrix, mention 2-dimensional arrays.
 
@@
 Here we discuss considerations that are subjective in nature, as they
-relate to our experience implementing all benchmarks in all
-languages. We need to take into account our experience using various
-languages prior to the beginning of this work:
+relate to individual experience implementing all benchmarks in all
+languages. Experience using various languages prior to the beginning
+of this work needs to be taken into account:
 \begin{itemize}
@@
 \end{itemize}
-To add to the subjectivity, each benchmark was first implemented in
+To add to the subjectivity, each benchmark is first implemented in
 Python then in other languages, with Rust typically being the last one
-to be implemented. Therefore the languages implemented last were
-supposedly easier, due to increased familiarity with the
+to be implemented. Therefore the languages implemented last represent
+supposedly easier tasks, due to increased familiarity with the
 algorithm. For all reasons mentioned above, measuring the
@@
 
-Using two threads instead of 1 improves performance by about 10\%,
-although it is unclear what the second thread is used for.
+% Using two threads instead of 1 improves performance by about 10\%,
+% although it is unclear what the second thread is used for.
 
@@
 
+\appendix
+
+\section{Summary of implementations}
+\label{app:impl}
+\begin{table}
+  \begin{tabular}{l|llll}
+    \toprule
+    Implementation & Language & Extra & Distance matrix representation
+    & Note \\
+    \midrule
+    \bottomrule   
+  \end{tabular}
+  \caption{Overview of all implementations}
+\end{table}
+
 \bibliographystyle{plain}

### 2021-02-18 paper/main.tex
paper progress

@@
 
-\title{Benchmark!!1}
+\title{A benchmark for the implementation of standard vehicle routing
+  algorithms}
 \author{
@@
 \section{The benchmarks}
+\label{sec:benchmark}
 We consider five different benchmark algorithms that perform tasks
@@
 
-For each benchmark the reference CPU time is provided by the c++14
+For each benchmark the reference CPU time is provided by the c++98
 implementation. We also develop a variant of c++14 using a nested
@@
 \subsection{Julia}
-Julia is a relatively recent programming language. Mention dates for
-first release (0.4?) as well as 1.0.
+Julia is a relatively recent programming language. It is a
+general-purpose language but it is also particularly aimed at
+computationally intensive tasks such as numerical
+analysis~\cite{Julia-2017}. Version 1.0 was 
+released in 2018 and the language is well-documented and has a strong
+community.
 
-Flat vs square matrix, mention 2-dimensional arrays.
+Compilation happens in a JIT fashion. Like Numpy but
+unlike all other programming languages considered, Julia supports
+multi-dimensional arrays, therefore a native two-dimensional array can be
+used to represent a distance matrix. We also develop an
+implementation using flat matrix representation in Julia, called
+\emph{julia-flat-matrix}; we will determine which version is better
+through experiment.
+
+One valuable feature of Julia is the fact that projects have an
+environment, including package dependencies with specific
+versions. The language makes it easy to \emph{instantiate} the
+project environment, i.e. recreate the same environment in which the
+project was developed, with the same packages in the same
+versions. This provides safety from new library versions which break
+backward compatibility.
 
 \subsection{Rust}
+Rust is also relatively recent, with version 1.0 released in 2015. It
+is a general-purpose language with an emphasis on memory safety and
+on performance. Notable software using Rust include Firefox, Dropbox
+and Cloudflare~\cite{rust-lang}.
+
+Programs in Rust are pre-compiled, like in C++ for
+instance. Like Julia, the language is well documented and has a strong
+community. Also like Julia, projects have an environment, and it is
+easy to recreate the environment for a given project, including
+dependencies and their specific version, which again provides safety
+from backward-compatibility-breaking library updates. The \emph{rust}
+implementation uses a flat matrix representation.
 
 \subsection{JavaScript}
-JavaScript is not expected to perform as well as, say, C++. However
-there are several reasons why it can be appealing to implement routing
-algorithms in JavaScript:
+JavaScript is likely one of the most popular programming languages in
+the World, as it is used on countless websites as well as mobile
+applications based on web technology. JavaScript is not expected to
+perform as well as, say, C++. However there are several reasons why it
+can be appealing to implement routing algorithms in JavaScript:
 \begin{itemize}
@@
 \item JavaScript engines have received considerable attention from
-  major companies and are the subject of fierce competition on
-  performance. As a result they have seen vast performance
-  improvements over the years and the trend is likely to continue.
+  major companies and been the subject of fierce competition in
+  relation with web browsers engines. As a result they have seen vast
+  performance improvements over the years and the trend is likely to continue.
 \end{itemize}
@@
 
-Flat vs nested.
+The \emph{javascript} implementation uses a flat matrix representation
+but there is also an implementation using nested matrix representation,
+called \emph{javascript-flat-matrix}.
 
@@
 
-Mention here compilers and compiler versions used.
+A summary of all versions used for compilers and interpreters is
+provided in Appendix~\ref{app:lang}.
+
+\subsection{Benchmark data}
+Each instance file contains one $n \times n$ distance matrix as well
+as $p$ randomly generated permutations, which are used as starting
+\emph{seeds} for the algorithms (e.g. as starting solution for 2-opt).
+For $n$ taking each value of $\{20, 40, 60\}$ and fixing $p = 1000$ we
+generate 40 instance files, for a total of 120 instance files. Each
+instance file contains 1000 permutations, therefore is run 1000
+different times for each benchmark. As mentioned in
+Section~\label{sec:benchmark} distances are Euclidean distances multiplied
+by 100 and truncated to their integer part. In the following each data
+point represents the total CPU time for running one implementation
+for each of the 1000 permutations of one given instance file.
 
@@
 
+\bibliographystyle{plain}
+\bibliography{benchmark}
+
 \appendix
@@
 
-\bibliographystyle{plain}
-\bibliography{benchmark}
+\section{Summary of language and compiler versions}
+\label{app:lang}
+\begin{table}
+  \begin{tabular}{l|l}
+    \toprule
+    Language/tool & language or compiler version & options \\
+    \midrule
+    C++14 & g++ & -Wall -ansi -pedantic -O3 -std=c++14 \\
+    C++98 & g++ & -Wall -ansi -pedantic -O3 -std=c++98 \\
+    Python & & \\
+    Pypy & & \\
+    Numpy & & \\
+    Numba & & \\
+    Java &  & \\
+    Julia & 1.5.3 & --check-bounds=no --inline=yes -O3 -t 1 \\
+    Rust & & \\
+    JavaScript & Node & \\
+    \bottomrule   
+  \end{tabular}
+  \caption{Overview of all implementations}
+\end{table}
 

### 2021-04-15 paper/main.tex
more description and comments

@@
 Purposes: determine which language we can tell our students to
-use to implement routing optimisation routines. Also determine which
+use to implement routing optimization routines. Also determine which
 language we can use ourselves.
@@
 \section{Experiments}
-All programs are run concurrently on an AMD Ryzen 5 3600 6-Core
-Processor at 3.6 GHz with 16 GB RAM running Linux. Twelve threads can
+All programs are run concurrently on an desktop computer with an AMD
+Ryzen 5 3600 6-Core Processor at 3.6 GHz with 16 GB RAM, running
+Linux. Twelve threads can
 run in parallel, however in our setting at most 5 threads are used at
@@
 
+\subsection{A comparison of Python implementations}
+
 \subsection{Impact of using a flat matrix}
+
 \subsection{Impact of using static arrays}
-\subsection{A comparison of Python implementations}
+
 \subsection{A comparison of C++ implementations}
 Also add a clang++ vs g++ comparison?
+
 \subsection{General cross-language comparison}
+
 \subsection{Cross-language comparison: fast languages}
@@
 We now look at the same benchmarks, using the same languages, but running
-in a different environment. We compare runs on an Intel i5-8265U
-4-core CPU at 1.6 GHz and on the previously mentioned AMD CPU. The
+in a different environment. We look at runs on an laptop Intel CPU
+(Intel i5-8265U, 4 cores, 1.6 GHz), on a Raspberry Pi 4 (ARMv7,
+4 cores, 1.5 GHz), on a shared cluster (Intel Xeon E5-2650 v3, 20
+cores, 3 GHz) and on the previously mentioned AMD CPU. The
 goal is not to observe which CPU provides the best performance, rather
 to determine whether the experimental observations made with one CPU
-are valid with the other. For this reason, we stick to reporting CPU
-times as a ratio to the CPU time required by the c++98 implementation.
-
-Perhaps also run on WU cluster and observe differences.
+are valid with the other. For this reason, we only report results that
+are different from the ones observed with the previous CPU.
+We note here that despite our best effort, language and compiler
+versions might vary between platforms, due to the diversity in
+package version availability on different systems. Ultimately, the
+goal of this analysis is not to obtain a very precise assessment of
+relative performance depending on compiler version and architecture, as it
+is of limited use to anyone using a slightly different architecture or
+compiler version; rather, we want to assess the reliability of
+previously observed results.
 
@@
 optimization algorithms is one that, by essence, does not have a final
-answer.
+answer. This being said, there is value in knowing how much
+performance is gained or lost when using a certain
+language on a certain architecture, or how much can be
+gained by using different data structures. The benchmark suite
+introduced in this paper is free software and can be used for that
+purpose. We encourage students and researchers in
+transport optimization to use it to support their implementation
+decisions. Further experimental analyses similar to the ones conducted
+here can also be good practice, for instance simply comparing the
+runtime performance of the same C++ code when compiled with different
+compilers, which is easier than it ever was.
+
+Additionally, some insight can already be taken straight out of the
+experiments conducted above. Using languages like Julia or Rust
+does not incur a significant runtime penalty when compared to more
+established languages like C++, and can actually be faster
+sometimes. Interpreted languages like Python or JavaScript nowadays
+have reasonably good runtime performance, under five times slower than
+pre-compiled languages, under the right conditions (using a JIT
+compiler). This means that the flexibility and ease of use of these
+languages can be worth the performance hit, depending on the
+application. Additionally, using JavaScript opens perspectives, as the
+integration of optimization algorithms in web applications is easier
+than it ever was, and can now be run client-side.
 
@@
     \midrule
-    C++14 & g++ & -Wall -ansi -pedantic -O3 -std=c++14 \\
-    C++98 & g++ & -Wall -ansi -pedantic -O3 -std=c++98 \\
-    Python & & \\
-    Pypy & & \\
-    Numpy & & \\
-    Numba & & \\
-    Java &  & \\
-    Julia & 1.5.3 & --check-bounds=no --inline=yes -O3 -t 1 \\
-    Rust & & \\
+    C++14 & g++ 10.2.1 & -Wall -ansi -pedantic -O3 -std=c++14 \\
+    C++98 & g++ 10.2.1 & -Wall -ansi -pedantic -O3 -std=c++98 \\
+    CPython & 3.9.2 & \\
+    Pypy & 7.3.3 & \\
+    % Numpy & & \\
+    % Numba & 0.53.1 & \\
+    Java & OpenJDK version 11 & \\
+    Julia & 1.6.0 & --check-bounds=no --inline=yes -O3 -t 1 \\
+    Rust & 1.51 & cargo build --release \\
     JavaScript & Node & \\
@@
 
+There are small variations in versions used on different
+platforms. Most notably, the Julia version used on the ARMv7 CPU is
+1.3, as it is the latest version available for that architecture.
+
 \end{document}

### 2021-04-15 scripts/compare.R
now the first command line argument is the directory to parse

@@ -5,7 +5,10 @@ library(ggplot2)
 
 ## reference implementation for ratio calculation
 refimpl <- 'c++14'
-csvdir <- 'csv/ryzen_5_3600'
+
+## result directory as first command line argument
+args = commandArgs(trailingOnly=TRUE)
+csvdir <- ifelse(length(args) ==0, 'results/ryzen_5_3600', args[1])
 
 ## first we glue all data together in one data frame
 fnames <- list.files(path=csvdir, pattern='*.csv')
